---
title: "Disease_data_preparation"
author: "Natasa Mortvanski"
date: "2023-04-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Disease data preprocessing

Loading necessary libraries:

```{r}
library(dplyr)
library(purrr)
library(stringr)
library(flextable)
#install.packages("here")
library(here)
library(tidyr)
```

## Load doanors metadata:
```{r}
hosp_donor_meta_all <- read.delim(here("00_raw_data/08_hospital_donor", "sample-metadata.tsv"))

hosp_donor_meta <- dplyr::select(hosp_donor_meta_all, SampleID, Peti)
names(hosp_donor_meta)[names(hosp_donor_meta) == 'SampleID'] <- 'sample_id'
names(hosp_donor_meta)[names(hosp_donor_meta) == 'Peti'] <- 'peti'

hosp_donor_meta <- na.omit(hosp_donor_meta) 

#Duplicated entries in the metadata file
hosp_donor_meta[duplicated(hosp_donor_meta$peti)==TRUE,]

hosp_donor_meta[hosp_donor_meta$peti=="226182837",]
hosp_donor_meta[hosp_donor_meta$peti=="226182873",]

#Eliminate problematic samples

hosp_donor_meta <- hosp_donor_meta %>%  filter(peti != "226182837" &
                                               peti != "226182873")

head(hosp_donor_meta)
```

## Load file connecting IDs of donors with SCFA data:

```{r}
SCFA_hosp_donor_ID <- read.delim(here("00_raw_data/08_hospital_donor/SCFA_taxonomy_analysis", "SCFA_normalised_IDS.csv"))
names(SCFA_hosp_donor_ID)[names(SCFA_hosp_donor_ID) == 'Donor'] <- 'donor'
names(SCFA_hosp_donor_ID)[names(SCFA_hosp_donor_ID) == 'Peti'] <- 'peti'
names(SCFA_hosp_donor_ID)[names(SCFA_hosp_donor_ID) == 'Donation.number'] <- 'donation_n'

SCFA_hosp_donor_ID <- SCFA_hosp_donor_ID %>% tidyr::unite("donor_donation", donor:donation_n, remove=FALSE)
SCFA_hosp_donor_ID <- subset(SCFA_hosp_donor_ID, select = -c(donor, donation_n))

SCFA_hosp_donor_ID <- SCFA_hosp_donor_ID %>% filter(peti != "226182837" & peti != "226182873")

head(SCFA_hosp_donor_ID)
```

##   Load SCFA data:
```{r}
SCFA_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor/SCFA_taxonomy_analysis", "SCFA_normalised.csv"))
names(SCFA_hosp_donor)[names(SCFA_hosp_donor) == 'Donor'] <- 'donor'
names(SCFA_hosp_donor)[names(SCFA_hosp_donor) == 'Donation.number'] <- 'donation_n'
names(SCFA_hosp_donor)[names(SCFA_hosp_donor) == 'Acetic.acid'] <- 'acetic_acid'
names(SCFA_hosp_donor)[names(SCFA_hosp_donor) == 'Propionic.acid'] <- 'propionic_acid'
names(SCFA_hosp_donor)[names(SCFA_hosp_donor) == 'Butyric.acid'] <- 'butyric_acid'
SCFA_hosp_donor <- SCFA_hosp_donor %>% tidyr::unite("donor_donation", donor:donation_n, remove=FALSE)
SCFA_hosp_donor <- subset(SCFA_hosp_donor, select = -c(donor, donation_n))

head(SCFA_hosp_donor)
```

## Merge hosp_donor_meta and SCFA_hosp_donor_ID

```{r}
IDs <- hosp_donor_meta %>% left_join(SCFA_hosp_donor_ID, by=c('peti' = 'peti'))

SCFA_hosp_donor_ID$peti[duplicated(SCFA_hosp_donor_ID$peti) ==TRUE]
hosp_donor_meta$peti[duplicated(hosp_donor_meta$peti)==TRUE]

IDs <- IDs[complete.cases(IDs),]

head(IDs)
```

## Check why IDs and SCFA_hosp_donor tables have different number of samples and fix text differences

```{r}
dplyr::anti_join(IDs, SCFA_hosp_donor, by = c("donor_donation" = "donor_donation"))
IDs <- IDs %>% mutate(donor_donation = str_replace(donor_donation, " _", "_"))
SCFA_hosp_donor <- SCFA_hosp_donor %>% mutate(donor_donation = str_replace(donor_donation, "_22th", "_22nd"))
SCFA_hosp_donor <- SCFA_hosp_donor %>% mutate(donor_donation = str_replace(donor_donation, "_21th", "_21st"))
IDs <- IDs %>% mutate(donor_donation = str_replace(donor_donation, "_21th", "_21st"))



#check after changes
dplyr::anti_join(IDs, SCFA_hosp_donor, by = c("donor_donation" = "donor_donation"))
dplyr::anti_join( SCFA_hosp_donor, IDs, by = c("donor_donation" = "donor_donation"))

```
Samples present in IDs table and not present in SCFA table:
GFG29M_3rd	

*JMJ44M_4th*	

AJL34M_2nd		
AJL34M_3rd		
AJL34M_4th		

Samples present in SCFA table and not present in IDs table:

*GFG29M_16th* (duplicated sample)
*SMP48F_2nd* (duplicated sample)
*JMJ44M_3rd*

CLA33F_12th
SPB34F_3rd
AGS29F_2.2nd  (in IDs there are two entries of AGS29F_2)
CGR32F_1st
RTF49M_9th




## Merge IDs and SCFA_hosp_donor
```{r}
SCFA_donors <- IDs %>% left_join(SCFA_hosp_donor, by=c('donor_donation' = 'donor_donation'))
SCFA_donors <- SCFA_donors[complete.cases(SCFA_donors),]

head(SCFA_donors)


# dplyr::anti_join(hosp_donor_meta_all, IDs, by = c("Peti" = "peti"))
```


Export the table of sample IDs to a text file for filtering the samples for QIIME2 analysis:
```{r}
write.table(SCFA_alpha_donors$sample_id, "~/Desktop/master_project/Data_Hospital_donors/sample_IDs_SCFA.txt", row.names=FALSE, quote=FALSE, col.names = FALSE)
```

Use this file to filter fastq files that have SCFA data by applying this bash script:

<!-- #!/bin/bash -->

<!-- mkdir ./SCFA_subset -->

<!-- # Define the input file -->
<!-- INFILE=~/Desktop/master_project/Data_Hospital_donors/sample_IDs_SCFA.txt -->

<!-- # Read the input file line by line -->
<!-- while read -r LINE -->
<!-- do -->
<!--     #NAME="$LINE" -->
<!--     printf '%s\n' "$LINE" -->
<!--     cp $LINE* ./SCFA_subset -->
<!-- done < "$INFILE" -->

Than execute sript "qiime2_analysis_DADA2_donors.sh" using files in SCFA_subset directory. 

## Import newly calculated alpha deversities

```{r}
#-----alpha diversity--------
shannon_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor/SCFA_taxonomy_analysis", "shannon_hosp_donor.tsv"))
chao1_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor/SCFA_taxonomy_analysis", "chao1_hosp_donor.tsv"))
fisher_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor/SCFA_taxonomy_analysis", "fisher_hosp_donor.tsv"))
margalef_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor/SCFA_taxonomy_analysis", "margalef_hosp_donor.tsv"))
simpson_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor/SCFA_taxonomy_analysis", "simpson_hosp_donor.tsv"))
gini_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor/SCFA_taxonomy_analysis", "gini_hosp_donor.tsv"))
menhinick_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor/SCFA_taxonomy_analysis", "menhinick_hosp_donor.tsv"))
strong_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor/SCFA_taxonomy_analysis", "strong_hosp_donor.tsv"))
faith_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor/SCFA_taxonomy_analysis", "faith_hosp_donor.tsv"))
pielou_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor/SCFA_taxonomy_analysis", "pielou_hosp_donor.tsv"))

all_hosp_donor_alpha <- list(shannon_hosp_donor, chao1_hosp_donor, fisher_hosp_donor, margalef_hosp_donor, gini_hosp_donor, menhinick_hosp_donor, strong_hosp_donor, simpson_hosp_donor, faith_hosp_donor, pielou_hosp_donor) 
all_hosp_donor_alpha <- all_hosp_donor_alpha %>% reduce(inner_join, by="X")
names(all_hosp_donor_alpha)[names(all_hosp_donor_alpha) == 'X'] <- 'sample_id'

#------------------------------
```

```{r}
SCFA_alpha_donors <- SCFA_donors %>% left_join(all_hosp_donor_alpha, by=c('sample_id' = 'sample_id'))

#Delete problematic sample
SCFA_alpha_donors <- SCFA_alpha_donors %>%  filter(donor_donation != 'SMP48F_2nd')

head(SCFA_alpha_donors)
```


## Export data frame with SCFA and alpha diversity data into .csv.gz file:

```{r}
write.table(SCFA_alpha_donors, file = here("01_tidy_data", "SCFA_alpha_donors.tsv"), row.names=FALSE, sep="\t", quote=FALSE)
```
