---
title: "Data preparation"
author:
- name: "Natasa Mortvanski"
  affiliation: Universitat Pompeu Fabra
  email: natasa.mortvanski01@estudiant.upf.edu
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document contains the code for data preparation (loading different alpha metrics, filtering of the samples and columns of interest) of American Gut Project data that is going to be used as [healthy control](#healthy-data-preprocessing) in this Master thesis, as well as data from various studies of different [pathological conditions](#disease-data-preprocessing) supposedly affecting gut microbiome alpha diversity.

# Healthy data preprocessing

Loading necessary libraries:

```{r}
library(dplyr)
#install.packages('purrr')
library(purrr)
library(tibble)
library(ggplot2)
library(stringr)
library(RColorBrewer)
library(gridExtra)
library(grid)
#install.packages("here")
library(here)
```

### About dataset:

As a source of samples for healthy control we chose American Gut Project study. Data and metadata are available on Qiita platform:

1. Subset of the American Gut Project dataset (healthy dataset) -> [dataset link](https://qiita.ucsd.edu/study/description/10317)

Loading the metadata file for AGP study samples:

```{r}
AGP_metadata <- read.delim(here("00_raw_data/01_AGP_data", "10317_20221012-092710.txt.gz"))

nrow(AGP_metadata)
```

Loading tables containing different alpha diversity metrics:

```{r}
shannon_AGP <- read.delim(here("00_raw_data/01_AGP_data", "shannon_AGP_all.tsv.gz"))
chao1_AGP <- read.delim(here("00_raw_data/01_AGP_data", "chao1_AGP_all.tsv.gz"))
menhinick_AGP <- read.delim(here("00_raw_data/01_AGP_data", "menhinick_AGP_all.tsv.gz"))
margalef_AGP <- read.delim(here("00_raw_data/01_AGP_data", "margalef_AGP_all.tsv.gz"))
simpson_AGP <- read.delim(here("00_raw_data/01_AGP_data", "simpson_AGP_all.tsv.gz"))
fisher_AGP <- read.delim(here("00_raw_data/01_AGP_data", "fisher_AGP_all.tsv.gz"))
gini_AGP <- read.delim(here("00_raw_data/01_AGP_data", "gini_AGP_all.tsv.gz"))
strong_AGP <- read.delim(here("00_raw_data/01_AGP_data", "strong_AGP_all.tsv.gz"))
faith_AGP <- read.delim(here("00_raw_data/01_AGP_data", "faith_AGP_all.tsv.gz"))
pielou_AGP <- read.delim(here("00_raw_data/01_AGP_data", "pielou_AGP_all.tsv.gz"))

# Make inverse Simpson
#simpson_AGP$inverse_simpson <- sapply(simpson_AGP$simpson, function(x) 1/x)

names(faith_AGP)[names(faith_AGP) == 'X.SampleID'] <- 'X'

# Merge all alpha diversity tables 
all_alpha <- list(shannon_AGP, chao1_AGP, menhinick_AGP, margalef_AGP, fisher_AGP, gini_AGP, strong_AGP, simpson_AGP, faith_AGP, pielou_AGP) 

all_alpha <- all_alpha %>% reduce(inner_join, by="X")
names(all_alpha)[names(all_alpha) == 'X'] <- 'sample_id'

nrow(all_alpha)
```

Totall number of samples in this dataset is `r nrow(all_alpha)`.

Now lets combine alpha metrics and metadata in the same dataframe (`AGP`):

```{r}
# Merge all_alpha and AGP_metadata
AGP <- all_alpha %>% left_join(AGP_metadata, by=c('sample_id' = 'sample_name'))

# Replace NA with empty string
AGP <- replace(AGP, is.na(AGP), "")

nrow(AGP)
```

We are left with `r nrow(AGP)` samples for which alpha metrics were calculated.
\
Lets filter out samples that are truly healthy:

```{r}
# Filter samples by age (only adult samples)
AGP_healthy_subset <- AGP %>%
  filter(age_cat== "20s" | 
         age_cat== "30s" | 
         age_cat== "40s" | 
         age_cat== "50s" | 
         age_cat== "60s")

# Filter out stool and feces samples
AGP_healthy_subset <- AGP_healthy_subset %>%
  filter(sample_type=="feces" | 
         sample_type=="Stool")

#Keep only samples with normal BMI
AGP_healthy_subset <- AGP_healthy_subset %>%
  filter(host_body_mass_index > 18.5 & 
         host_body_mass_index < 25)

# Filtering criteria No. 1
## Other solution is to apply only filters used in original AGP study:
AGP_healthy_subset_1 <- AGP_healthy_subset %>%
  filter(antibiotic_history=="I have not taken antibiotics in the past year." & 
         diabetes=="I do not have this condition" & 
         ibd=="I do not have this condition")

# Filtering criteria No. 2
## AGP filters + no IBS or Cdiff
AGP_healthy_subset_2 <- AGP_healthy_subset %>%
  filter(antibiotic_history=="I have not taken antibiotics in the past year." & 
         diabetes=="I do not have this condition" & 
         ibd=="I do not have this condition" &
         ibs=="I do not have this condition" &
         cdiff=="I do not have this condition")

# Filtering criteria No. 3
## Include only samples that have reported not having any of the following diseases
AGP_healthy_subset_3 <- AGP_healthy_subset %>%
  filter(antibiotic_history=="I have not taken antibiotics in the past year." & 
         diabetes=="I do not have this condition" & 
         ibd=="I do not have this condition" &
         ibs=="I do not have this condition" &
         cdiff=="I do not have this condition" & 
         cancer=="I do not have this condition" &
         mental_illness=="No" &
         mental_illness_type_substance_abuse=="false" &
         mental_illness_type_depression=="false" &
         asd=="I do not have this condition" &
         alzheimers=="I do not have this condition" &
         add_adhd=="I do not have this condition" &
         epilepsy_or_seizure_disorder=="I do not have this condition" &
         sibo=="I do not have this condition" &   # small intestinal bacterial overgrowth
         pku =="I do not have this condition" &   # phenylketonuria
         thyroid=="I do not have this condition" &
         kidney_disease=="I do not have this condition" &
         clinical_condition=="I do not have this condition" &
         acid_reflux=="I do not have this condition" &
         autoimmune=="I do not have this condition" &
         cardiovascular_disease=="I do not have this condition" &
         lung_disease =="I do not have this condition" &
         liver_disease =="I do not have this condition")

# Number of samples without any reported clinical condition
nrow(AGP_healthy_subset_1)  #2292

# Number of samples after filtering only by criteria reported by AGP paper
nrow(AGP_healthy_subset_2)  #1465

# Number of samples after filtering only by criteria reported by AGP paper + without IBS and Cdiff
nrow(AGP_healthy_subset_3)  #600
```
First subset of data filtered only by criteria from AGP paper contains `r nrow(AGP_healthy_subset_1)` samples.
Second subset of data filtered by criteria from paper with additionally filtered out IBS and Cdiff positive samples contain `r nrow(AGP_healthy_subset_2)` samples.
Finally, the third subset that consists of samples with none reported diseases contains `r nrow(AGP_healthy_subset_3)` samples.

Lets see if there is a difference in values of alpha diversity metrics in different subsets:

```{r}
# make tables of summaries of all 4 subsets of AGP data
tabs <- list()
subsets <- list(AGP_healthy_subset, AGP_healthy_subset_1, AGP_healthy_subset_2, AGP_healthy_subset_3)

for(x in 1:length(subsets)){
tabs[[x]] <- subsets[[x]] %>% 
  summarise(shannon_entropy,
            chao1,
            menhinick,
            margalef,
            simpson,
            pielou_evenness,
            #inverse_simpson,
            fisher_alpha,
            gini_index,
            strong,
            faith_pd) %>% 
  broom::tidy()
}

# Extract minimums, means and maximums of each metric and combine them in new tables
cols <- c("column", "mean", "min", "max")
metric <- c("shannon_entropy", "chao1", "menhinick", "margalef", "fisher_alpha", "simpson", "gini_index", "strong", "pielou_evenness", "faith_pd" )
alpha_tables <- list()

for (i in 1:length(metric)){
  alpha_tables[[i]] <- tibble()
  for (x in tabs){
    alpha_tables[[i]] <- bind_rows(alpha_tables[[i]], filter(x, column == metric[[i]]))
  }
  alpha_tables[[i]] <- alpha_tables[[i]][, names(alpha_tables[[i]]) %in% cols]
  alpha_tables[[i]] <- alpha_tables[[i]] %>% add_column(subsample = c("AGP fecal samples of participants aged 20-69", "healthy AGP participants without any reported disease", "healthy AGP participants (AGP paper criteria)", "healthy AGP participants (AGP paper criteria + no IBD nor Cdiff)"), .before = "column") 
}

all_alpha_tables<- tibble()
for(x in 1:length(alpha_tables)){
  all_alpha_tables <- bind_rows(all_alpha_tables, alpha_tables[[x]])
}
```

```{r}
# Plot values of min, mean and max for each subset of AGP data
all_alpha_tables %>%
  ggplot(aes(x= min, y=subsample)) + 
  geom_dotplot(binaxis='y', stackdir='center') +
  theme(axis.text.x = element_text(angle = 35, vjust = 1, hjust=1)) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 35))+
  facet_wrap(column ~ ., scales = "free_x", ncol=5)

all_alpha_tables %>%
  ggplot(aes(x= mean, y=subsample)) + 
  geom_dotplot(binaxis='y', stackdir='center') +
  theme(axis.text.x = element_text(angle = 35, vjust = 1, hjust=1)) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 35))+
  facet_wrap(column ~ ., scales = "free_x", ncol=5)

all_alpha_tables %>%
  ggplot(aes(x= max, y=subsample)) + 
  geom_dotplot(binaxis='y', stackdir='center') +
  theme(axis.text.x = element_text(angle = 35, vjust = 1, hjust=1)) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 35))+
  facet_wrap(column ~ ., scales = "free_x", ncol=5) 

```

Group of healthy AGP participants without any reported disease has the highest 
minimums and means of all alpha diversity indexes except from Gini's and Strong's 
that have the lowest mean.

Lets explore how distributions of different metrics change in different subsets:

Generate a vector of 10 random colors for histograms:

```{r}
qual_col_pals <- brewer.pal.info[brewer.pal.info$category == 'qual',]
colors <- unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
```

```{r}
library(purrr)
subset_list <- list(AGP_healthy_subset, AGP_healthy_subset_1, AGP_healthy_subset_2, AGP_healthy_subset_3)
subset_names <- c("AGP_healthy_subset", "AGP_healthy_subset_1", "AGP_healthy_subset_2", "AGP_healthy_subset_3")
subset_list <- setNames(subset_list, subset_names)

histo <- vector('list', length(subset_list)*length(metric))

for (i in 1:length(metric)){
  for (j in 1:length(subset_list)){
    histo[[i]][[j]] <- subset_list[[j]] %>% ggplot(aes(x = .data[[metric[i]]])) +
      geom_histogram(aes(y=..density..), colour="black", fill="white", bins=30)+
      geom_density(alpha=.2, fill=colors[i]) +
      xlab(label = paste(names(subset_list[j]),"(n =", nrow(subset_list[[j]]), ")", sep=' ')) + 
      ylab(label ="density")
  }
}

plot_list <- list()
for (i in 1:length(metric)){
  plot_list[[i]] <- list()
  for (j in 1:length(subset_list)){
    plot_list[[i]][[j]] <- histo[[i]][[j]]
  }
}


for (i in 1:length(metric)){
  grid.arrange(grobs = plot_list[[i]], ncol=2, top = textGrob(paste ("Density distributions of data in different subsets (different filtering criteria) for alpha metric:", metric[i], sep =" "), gp=gpar(fontsize=10,font=2)))
}
```

For now lets use the subset of healthy samples after most restrictive filtering (healthy AGP participants without any reported disease):

```{r}
AGP_healthy_subset <- AGP_healthy_subset_3

nrow(AGP_healthy_subset)
```

There is `r nrow(AGP_healthy_subset)` healthy samples.

One of the things we are interested in is the nutrition of study participants.
AGP supposedly incorporates data from Vioscreen questionaire about nutrition habits.
Let's examine those columns:

```{r}
# Examine Vioscreen data columns
cols <- grep("^vioscreen.?", names(AGP_healthy_subset), value=T)

# Show unique values in each of the vioscreen columns
#lapply(AGP_healthy_subset[,cols], unique)
```

Vioscreen columns are highly redundand (empty). Furthermore, only available 
values for vioscreen columns are: "" and "not applicable". We can delete 
vioscreen columns from the table AGP_healthy_subset because they are not 
informative:

```{r}
# Delete vioscreen columns
AGP_healthy_subset <- AGP_healthy_subset[, !names(AGP_healthy_subset) %in% cols]
```

Delete columns that are not informative enough, for example mostly empty, mostly 
consisting of 'Unspecified' answer or uniform (only have one category, so useless 
for comparisons):

```{r}
# Delete almost empty columns (missing more than 90% of the samples)
empty_columns <- colSums(is.na(AGP_healthy_subset) | AGP_healthy_subset == "") > nrow(AGP_healthy_subset)*0.9
AGP_healthy_subset <- AGP_healthy_subset[, !names(AGP_healthy_subset) %in% empty_columns]

# Delete columns that have more than 80% Uncpecified answers
unspec_columns <- colSums(AGP_healthy_subset == "Unspecified" | AGP_healthy_subset == "") > nrow(AGP_healthy_subset)*0.8
AGP_healthy_subset <- AGP_healthy_subset[, !unspec_columns]

#Delete columns with uniform values (all values are the same)
uniform_columns <- vapply(AGP_healthy_subset, function(x) length(unique(x)) == 1, logical(1L))
AGP_healthy_subset <- AGP_healthy_subset[, !uniform_columns]
```

Furthermore, we are not interested in examining effects of Covid or surfing on 
gut microbiome so we are going to delete these columns:

```{r}
# Delete covid related columns
cols <- grep("^covid?", names(AGP_healthy_subset), value=T)
AGP_healthy_subset <- AGP_healthy_subset[, !names(AGP_healthy_subset) %in% cols]

# Delete columns related to surfing data
cols <- grep("^surf_.?", names(AGP_healthy_subset), value=T)
AGP_healthy_subset <- AGP_healthy_subset[, !names(AGP_healthy_subset) %in% cols]
```

Other redundand columns:

* host_age_units - in years for all samples
* host_age_normalized_years is the same as host_age, 
* host_subject_id and anonymized_name - we already have sample identifiers 
* (sample_name), we don't need subject ID since we don't need to identify subjects 
* roommates_in_study - irrelevant since we can't identify them, 
* toilet_water_access - irrelevant
* collection_timestamp - irrelevant

```{r}
# Delete columns
cols <- c('host_age_units', 'host_age_normalized_years', 'host_subject_id', 'roommates_in_study', 'toilet_water_access', 'anonymized_name', 'collection_timestamp')
AGP_healthy_subset <- AGP_healthy_subset[, !names(AGP_healthy_subset) %in% cols]

ncol(AGP_healthy_subset) 

#colnames(AGP_healthy_subset)
```

We are left with `r ncol(AGP_healthy_subset)` columns relating to participants' 
nutrition habits, allergies, physical activity, cohabitation etc.

Check other columns of interest
```{r}
table(AGP_healthy_subset$sex)
# cathegories: female, male, other, unspecified

# Delete other and unspecified
AGP_healthy_subset <- AGP_healthy_subset %>%
  filter(sex=="female" | sex=="male")

# table(AGP_healthy_subset$bmi_cat)
# # cathegories: Normal, Unspecified  
# 
# # Delete Unspecified
# AGP_healthy_subset <- AGP_healthy_subset %>%
#   filter(bmi_cat !="Unspecified")
```

```{r}
nrow(AGP)
nrow(AGP_healthy_subset)

table(AGP_healthy_subset$bmi_cat)
```

Export table into csv format:

```{r}
write.csv(AGP, gzfile(here("01_tidy_data", "AGP_all.csv.gz")), row.names=FALSE)
write.csv(AGP_healthy_subset, gzfile(here("01_tidy_data", "AGP_healthy.csv.gz")), row.names=FALSE)
```

# Disease data preprocessing

### About datasets

Following studies will be used as source of samples (from American donors) of different pathological conditions that affect alpha diversity:

[1.](#ibd-datasets) Multi-omics of the gut microbial ecosystem in inflammatory bowel diseases, [dataset link](https://qiita.ucsd.edu/study/description/11484) and Metaomics Reveals Microbiome Based Proteolysis as a Driver of Ulcerative Colitis Severity, [dataset link](https://qiita.ucsd.edu/study/description/11549)

[2.](#longitudinal-chron's-disease-study) Guiding longitudinal sampling in inflammatory bowel diseases cohorts -> [dataset link](https://qiita.ucsd.edu/study/description/2538#)

[3.](#changes-following-fecal-microbial-transplantation-for-recurrent-cdi) Dynamic changes in short- and long-term bacterial composition following fecal microbiota transplantation for recurrent Clostridium difficile infection -> [dataset link](https://qiita.ucsd.edu/study/description/1924)

[4.](#fecal-transplantation-ibd-and-cdi-dataset) Changes in Microbial Ecology after Fecal Microbiota Transplantation for recurrent C. difficile Infection Affected by Underlying Inflammatory Bowel Disease -> [dataset link](https://qiita.ucsd.edu/study/description/10057)

[5.](#hospital-clinic's-cdi-samples) Dataset of Hospital Clinic's CDI patients' samples pre and post fecal microbiom transplantation 

All studies' data and metadata is available on Qiita platform.

```{r}
# Prepare metadata files

# for study 1.
IBD_meta_1 <- read.delim(here("00_raw_data/02_IBD", "sample_information_from_prep_4068.tsv.gz"))  

# for study 2.
IBD_meta_2 <- read.delim(here("00_raw_data/02_IBD", "sample_information_from_prep_8715.tsv.gz"))

# for study 3.
CD_2a <- read.delim(here("00_raw_data/03_CD_longitudinal", "sample_information_from_prep_569.tsv.gz"))

# for study 4.
C_diff_tr <- read.delim(here("00_raw_data/04_C_difficile", "sample_information_from_prep_262.tsv.gz"))

# for study 5.
trans_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "sample_information_transpl_CDI_IBD.txt.gz"))

# for hospital's CDI dataset 
hosp_CDI_meta_all <- read.delim(here("00_raw_data/06_hospital_CDI", "sample-metadata.tsv.gz"))

# change the name of second column in IBD_meta_1 so it matches colimn names n IBD_meta_2
names(IBD_meta_1)[names(IBD_meta_1) == 'diagnosis'] <- 'condition'
names(IBD_meta_2)[names(IBD_meta_2) == 'disease'] <- 'condition'
names(trans_IBD_CDI)[names(trans_IBD_CDI) == 'pathology'] <- 'condition'
names(hosp_CDI_meta_all)[names(hosp_CDI_meta_all) == 'Sample.ID'] <- 'sample_id'
```

## IBD datasets

Lets upload first two studies related to Inflammatory bowel disease (Ulcerative colitis and Crohn's disease):

```{r}
IBD_meta_1 <- select(IBD_meta_1, sample_id, sex, race, consent_age, age_at_diagnosis, condition, qiita_study_id)

IBD_meta_2 <- select(IBD_meta_2, sample_id, sex, race, age, age_diagnosis, condition, qiita_study_id)

names(IBD_meta_1)[names(IBD_meta_1) == 'consent_age'] <- 'age'
names(IBD_meta_2)[names(IBD_meta_2) == 'age_diagnosis'] <- 'age_at_diagnosis'

IBD_meta_1[IBD_meta_1 == "White"] <-'Caucasian'

IBD_meta_2[IBD_meta_2 == "F"] <- 'female'
IBD_meta_2[IBD_meta_2 == "M"] <- 'male'
```

```{r}
# merge them in single dataframe
IBD_meta_all <-  rbind(IBD_meta_1, IBD_meta_2)
```

Filtering:

```{r}
# Filter for age: 20-69
IBD_meta_all <- IBD_meta_all %>%
  filter(age > 20 & age < 69)
```

There is no info about BMI in neither of these two datasets' metadata.

Load calculated alpha metrics for thiese two datasets:
```{r}
# Load .tsv.gz files from inflammatory bowel disease analysis
shannon_IBD <- read.delim(here("00_raw_data/02_IBD", "shannon_IBD.tsv.gz"))
chao1_IBD <- read.delim(here("00_raw_data/02_IBD", "chao1_IBD.tsv.gz"))
fisher_IBD <- read.delim(here("00_raw_data/02_IBD", "fisher_IBD.tsv.gz"))
margalef_IBD <- read.delim(here("00_raw_data/02_IBD", "margalef_IBD.tsv.gz"))
simpson_IBD <- read.delim(here("00_raw_data/02_IBD", "simpson_IBD.tsv.gz"))
gini_IBD <- read.delim(here("00_raw_data/02_IBD", "gini_IBD.tsv.gz"))
menhinick_IBD <- read.delim(here("00_raw_data/02_IBD", "menhinick_IBD.tsv.gz"))
strong_IBD <- read.delim(here("00_raw_data/02_IBD", "strong_IBD.tsv.gz"))
faith_IBD <- read.delim(here("00_raw_data/02_IBD", "faith_IBD.tsv.gz"))
pielou_IBD <- read.delim(here("00_raw_data/02_IBD", "pielou_IBD.tsv.gz"))

#simpson_IBD$inverse_simpson <- sapply(simpson_IBD$simpson, function(x) 1/x)

names(faith_IBD)[names(faith_IBD) == 'X.SampleID'] <- 'X'

#put all data frames into list
all_IBD <- list(shannon_IBD, chao1_IBD, menhinick_IBD, margalef_IBD, fisher_IBD, gini_IBD, strong_IBD, simpson_IBD, faith_IBD, pielou_IBD) 

#merge all data frames together
all_IBD <- all_IBD %>% reduce(inner_join, by="X")

names(all_IBD)[names(all_IBD) == 'X'] <- 'sample_id'

all_IBD$sample_id <- str_replace(all_IBD$sample_id, "^.*?\\.", "")
```

Merge with metadata:

```{r}
all_IBD <- all_IBD %>% left_join(IBD_meta_all, by=c('sample_id' = 'sample_id'))

all_IBD <- all_IBD[complete.cases(all_IBD),]
```

```{r}
nrow(all_IBD)
```

After filtering we are left with `r nrow(all_IBD)` samples for which we have calculated all chosen alpha metrics.

Export data frame into .csv.gz file
```{r}
write.csv(all_IBD, gzfile(here("01_tidy_data", "IBD_all.csv.gz")), row.names=FALSE)
```

## Longitudinal Chron's disease study

Lets upload data from the longitudinal study related to Crohn's disease:

```{r}
CD_2 <- select(CD_2a, sample_name, sex, age, description, sample_type, weeks, env_package, ibd, surgery_and_ibd, surgery_type, qiita_study_id)

names(CD_2)[names(CD_2) == 'sample_name'] <- 'sample_id'
names(CD_2)[names(CD_2) == 'ibd'] <- 'condition'


# Load .tsv.gz files from inflammatory bowel disease analysis
shannon_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_shannon.tsv.gz"))
chao1_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_chao1.tsv.gz"))
fisher_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_fisher.tsv.gz"))
margalef_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_margalef.tsv.gz"))
simpson_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_simpson.tsv.gz"))
gini_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_gini.tsv.gz"))
menhinick_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_menhinick.tsv.gz"))
strong_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_strong.tsv.gz"))
faith_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_faith.tsv.gz"))
pielou_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_pielou.tsv.gz"))

#simpson_CD_2$inverse_simpson <- sapply(simpson_CD_2$simpson, function(x) 1/x)

names(faith_CD_2)[names(faith_CD_2) == 'X.SampleID'] <- 'X'

#put all data frames into list
all_CD_2 <- list(shannon_CD_2, chao1_CD_2, fisher_CD_2, margalef_CD_2, gini_CD_2, menhinick_CD_2, strong_CD_2, simpson_CD_2, faith_CD_2, pielou_CD_2)

#merge all data frames together
all_CD_2 <- all_CD_2 %>% reduce(inner_join, by="X")

names(all_CD_2)[names(all_CD_2) == 'X'] <- 'sample_id'

# match sample_id from table ... and IBD_meta_all and add disease status from 
# table IBD_meta_all to corresponding sample

all_CD_2 <- all_CD_2 %>% left_join(CD_2, by=c('sample_id' = 'sample_id'))

all_CD_2 <- all_CD_2[complete.cases(all_CD_2),]
```

Filtering:

```{r}
# Filter for age and BMI
all_CD_2 <- all_CD_2 %>%
  filter(age > 20 & age < 69)

all_CD_2 <- all_CD_2 %>%
  filter(sample_type == "feces")
```

There is no information about BMI in this dataset's metadata.

```{r}
table(all_CD_2$description, all_CD_2$condition)

nrow(all_CD_2)
```

After filtering we are left with `r nrow(all_CD_2)` samples for which we have calculated all chosen alpha metrics.

Export data frame into .tsv.gz file:

```{r}
write.csv(all_CD_2, gzfile(here("01_tidy_data", "all_CD_2.csv.gz")), row.names=FALSE)
```

# Changes following fecal microbial transplantation for recurrent CDI

Lets load data from study about short- and long-term microbiome changes following FMT for recurrent C. difficile infection:

```{r}
C_diff_trans <- select(C_diff_tr, sample_name, animations_subject, sex, race, host_age, host_body_mass_index, day_relative_to_fmt, disease_state, qiita_study_id)

names(C_diff_trans)[names(C_diff_trans) == 'sample_name'] <- 'sample_id'
names(C_diff_trans)[names(C_diff_trans) == 'host_age'] <- 'age'
names(C_diff_trans)[names(C_diff_trans) == 'host_body_mass_index'] <- 'body_mass_index'
names(C_diff_trans)[names(C_diff_trans) == 'day_relative_to_fmt'] <- 'day_since_fmt'


# Load .tsv.gz files from inflammatory bowel disease analysis
shannon_C_diff_trans <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_shannon.tsv.gz"))
chao1_C_diff_trans <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_chao1.tsv.gz"))
fisher_C_diff_trans <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_fisher.tsv.gz"))
margalef_C_diff_trans <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_margalef.tsv.gz"))
simpson_C_diff_trans <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_simpson.tsv.gz"))
gini_C_diff_trans <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_gini.tsv.gz"))
menhinick_C_diff_trans <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_menhinick.tsv.gz"))
strong_C_diff_trans <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_strong.tsv.gz"))
pielou_C_diff_trans <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_pielou.tsv.gz"))
faith_C_diff_trans <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_faith.tsv.gz"))

names(faith_C_diff_trans)[names(faith_C_diff_trans) == 'X.SampleID'] <- 'X'

#simpson_C_diff_trans$inverse_simpson <- sapply(simpson_C_diff_trans$simpson, function(x) 1/x)

#put all data frames into list
all_C_diff_trans <- list(shannon_C_diff_trans, chao1_C_diff_trans, fisher_C_diff_trans, margalef_C_diff_trans, gini_C_diff_trans, menhinick_C_diff_trans, strong_C_diff_trans, faith_C_diff_trans, simpson_C_diff_trans, pielou_C_diff_trans)   

#merge all data frames together
all_C_diff_trans <- all_C_diff_trans %>% reduce(inner_join, by="X")

names(all_C_diff_trans)[names(all_C_diff_trans) == 'X'] <- 'sample_id'

#join C_diff_trans metadata with alpha metrics

all_C_diff_trans <- all_C_diff_trans %>% left_join(C_diff_trans, by=c('sample_id' = 'sample_id'))

#keep only complete cases

all_C_diff_trans <- all_C_diff_trans[complete.cases(all_C_diff_trans),]
nrow(all_C_diff_trans)
```

Filtering:

```{r}
all_C_diff_trans <- all_C_diff_trans %>%
  filter(age > 20 & age < 69)

all_C_diff_trans <- all_C_diff_trans %>%
  filter(animations_subject != "Patient" & 
         animations_subject != "Donor")
```

```{r}
table(C_diff_tr$host_body_mass_index, C_diff_tr$animations_subject)
```

Subject CD1 has a BMI of 29.3;
Subject CD2 has a BMI of 27.1;
Subject CD3 has a BMI of 32.5;
Subject CD4 has a BMI of 23.9.

All except CD4 have BMI over 25 which is considered owerveight.
For purpose of comparing improvement after FMT in these four subjects, we are not going to filter out based on BMI.
Furthermore, this is not necessary because we wont compare these samples with healthy dataset.

```{r}
nrow(all_C_diff_trans)
```

After filtering we are left with `r nrow(all_C_diff_trans)` samples for which we have calculated all chosen alpha metrics.

Export data frame into .tsv.gz file:
```{r}
write.csv(all_C_diff_trans, gzfile(here("01_tidy_data", "all_C_diff_trans.csv.gz")), row.names=FALSE)
```

Exported data frames are stored in 01_tidy_data directory of this project. They are going to be used for further statistical analysis.

## Fecal transplantation IBD and CDI dataset

Finally, lets load data from the study about effects of FMT (Fecal Microbiota Transplantation) for recurrent C. difficile infection with underlying inflammatory bowel disease:

```{r}
trans_IBD_CDI <- select(trans_IBD_CDI, sample_name, animations_subject, sex, age, body_mass_index, condition, day_since_fmt, donor_or_patient, qiita_study_id)

names(trans_IBD_CDI)[names(trans_IBD_CDI) == 'sample_name'] <- 'sample_id'

trans_IBD_CDI[trans_IBD_CDI == "Crohn's"] <-'CDI + CD'
trans_IBD_CDI[trans_IBD_CDI == "microscopic colitis"] <-'CDI + MC'
trans_IBD_CDI[trans_IBD_CDI == "Ulcerative colitis"] <-'CDI + UC'
trans_IBD_CDI[trans_IBD_CDI == "no IBD"] <-'CDI'
trans_IBD_CDI$condition[trans_IBD_CDI$condition == "Not applicable"] <-'Donors'

# Load .tsv.gz files from inflammatory bowel disease analysis
shannon_trans_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_shannon.tsv.gz"))
chao1_trans_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_chao1.tsv.gz"))
fisher_trans_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_fisher.tsv.gz"))
margalef_trans_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_margalef.tsv.gz"))
simpson_trans_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_simpson.tsv.gz"))
gini_trans_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_gini.tsv.gz"))
menhinick_trans_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_menhinick.tsv.gz"))
strong_trans_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_strong.tsv.gz"))
faith_trans_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_faith.tsv.gz"))
pielou_trans_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_pielou.tsv.gz"))

#simpson_trans_IBD_CDI$inverse_simpson <- sapply(simpson_trans_IBD_CDI$simpson, function(x) 1/x)

names(faith_trans_IBD_CDI)[names(faith_trans_IBD_CDI) == 'X.SampleID'] <- 'X'

#put all data frames into list
all_trans_IBD_CDI <- list(shannon_trans_IBD_CDI, chao1_trans_IBD_CDI, fisher_trans_IBD_CDI, margalef_trans_IBD_CDI, gini_trans_IBD_CDI, menhinick_trans_IBD_CDI, strong_trans_IBD_CDI, simpson_trans_IBD_CDI, faith_trans_IBD_CDI, pielou_trans_IBD_CDI) 

#merge all data frames together
all_trans_IBD_CDI <- all_trans_IBD_CDI %>% reduce(inner_join, by="X")

names(all_trans_IBD_CDI)[names(all_trans_IBD_CDI) == 'X'] <- 'sample_id'

all_trans_IBD_CDI <- all_trans_IBD_CDI %>% left_join(trans_IBD_CDI, by=c('sample_id' = 'sample_id'))

all_trans_IBD_CDI <- all_trans_IBD_CDI[complete.cases(all_trans_IBD_CDI),]
```

Filtering:

```{r}
all_trans_IBD_CDI <- all_trans_IBD_CDI %>%
  filter(age > 20 & age < 69)

all_trans_IBD_CDI_donors <- all_trans_IBD_CDI %>%
  filter(animations_subject == "Donors (CD and IBD)")

all_trans_IBD_CDI <- all_trans_IBD_CDI %>%
   filter(body_mass_index > 18.5 &
          body_mass_index < 25)

all_trans_IBD_CDI <- rbind(all_trans_IBD_CDI, all_trans_IBD_CDI_donors)

nrow(all_trans_IBD_CDI)
```

After filtering we are left with `r nrow(all_trans_IBD_CDI)` samples for which we have calculated all chosen alpha metrics.

Export data frame into .tsv.gz file:

```{r}
write.csv(all_trans_IBD_CDI, gzfile(here("01_tidy_data", "all_trans_IBD_CDI.csv.gz")), row.names=FALSE)
```

## Hospital dataset - CDI samples

```{r}
hosp_CDI_meta <- select(hosp_CDI_meta_all, sample_id, sample.origin, sex, age, time, CDItreatment.PREFMT, severeICD, Clinicalrecurrence)
names(hosp_CDI_meta)[names(hosp_CDI_meta) == 'sample.origin'] <- 'condition'
names(hosp_CDI_meta)[names(hosp_CDI_meta) == 'time'] <- 'day_since_fmt'
names(hosp_CDI_meta)[names(hosp_CDI_meta) == 'CDItreatment.PREFMT'] <- 'antibiotic_use'
names(hosp_CDI_meta)[names(hosp_CDI_meta) == 'Clinicalrecurrence'] <- 'clinical_recurrence'

# Load .tsv.gz files from inflammatory bowel disease analysis
shannon_hosp_CDI <- read.delim(here("00_raw_data/06_hospital_CDI", "shannon_hosp_CDI.tsv.gz"))
chao1_hosp_CDI <- read.delim(here("00_raw_data/06_hospital_CDI", "chao1_hosp_CDI.tsv.gz"))
fisher_hosp_CDI <- read.delim(here("00_raw_data/06_hospital_CDI", "fisher_hosp_CDI.tsv.gz"))
margalef_hosp_CDI <- read.delim(here("00_raw_data/06_hospital_CDI", "margalef_hosp_CDI.tsv.gz"))
simpson_hosp_CDI <- read.delim(here("00_raw_data/06_hospital_CDI", "simpson_hosp_CDI.tsv.gz"))
gini_hosp_CDI <- read.delim(here("00_raw_data/06_hospital_CDI", "gini_hosp_CDI.tsv.gz"))
menhinick_hosp_CDI <- read.delim(here("00_raw_data/06_hospital_CDI", "menhinick_hosp_CDI.tsv.gz"))
strong_hosp_CDI <- read.delim(here("00_raw_data/06_hospital_CDI", "strong_hosp_CDI.tsv.gz"))
faith_hosp_CDI <- read.delim(here("00_raw_data/06_hospital_CDI", "faith_hosp_CDI.tsv.gz"))
pielou_hosp_CDI <- read.delim(here("00_raw_data/06_hospital_CDI", "pielou_hosp_CDI.tsv.gz"))

#put all data frames into list
all_hosp_CDI <- list(shannon_hosp_CDI, chao1_hosp_CDI, fisher_hosp_CDI, margalef_hosp_CDI, gini_hosp_CDI, menhinick_hosp_CDI, strong_hosp_CDI, simpson_hosp_CDI, faith_hosp_CDI, pielou_hosp_CDI) 

#merge all data frames together
all_hosp_CDI <- all_hosp_CDI %>% reduce(inner_join, by="X")

names(all_hosp_CDI)[names(all_hosp_CDI) == 'X'] <- 'sample_id'

all_hosp_CDI <- all_hosp_CDI %>% left_join(hosp_CDI_meta, by=c('sample_id' = 'sample_id'))

# This filters out pool samples because of the lack of entries in metadata
all_hosp_CDI <- all_hosp_CDI[complete.cases(all_hosp_CDI),]
```

Filtering:

```{r}
all_hosp_CDI <- all_hosp_CDI %>%
  filter(age > 20 & age < 69)

all_hosp_CDI <- all_hosp_CDI %>%
  filter(condition == "Cdif")

nrow(all_hosp_CDI)
```
We have `r nrow(all_hosp_CDI)` samples for which we have calculated all chosen alpha metrics.

Export data frame into .tsv.gz file:

```{r}
write.csv(all_hosp_CDI, gzfile(here("01_tidy_data", "all_hosp_CDI.csv.gz")), row.names=FALSE)
```

# Overview of descriptive statistics of different data sets:

```{r}
data.frame(AGP = nrow(AGP_healthy_subset), IBD = nrow(all_IBD), CD2 = nrow(all_CD_2), C.diff = nrow(all_C_diff_trans), FMT_CDI_IBD = nrow(all_trans_IBD_CDI))
```

# Session information

```{r}
sessionInfo()
```

# References
