---
title: "Disease_data_preparation"
author: "Natasa Mortvanski"
date: "2023-04-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Disease data preprocessing

Loading necessary libraries:

```{r}
library(dplyr)
#install.packages('purrr')
library(purrr)
library(tibble)
library(ggplot2)
library(stringr)
library(RColorBrewer)
library(gridExtra)
library(grid)
#install.packages("here")
library(here)
```

### About datasets

Following studies will be used as source of samples (from American donors) of different pathological conditions that affect alpha diversity:

[1.](#ibd-datasets) Multi-omics of the gut microbial ecosystem in inflammatory bowel diseases, [dataset link](https://qiita.ucsd.edu/study/description/11484) 

[2.](#ibd-datasets) Metaomics Reveals Microbiome Based Proteolysis as a Driver of Ulcerative Colitis Severity, [dataset link](https://qiita.ucsd.edu/study/description/11549)

[3.](#longitudinal-chron's-disease-study) Guiding longitudinal sampling in inflammatory bowel diseases cohorts -> [dataset link](https://qiita.ucsd.edu/study/description/2538#)

[4.](#cdi-samples) Gut microbiome predictors of treatment response and recurrence in primary Clostridium difficile infection -> study accession: PRJNA342347

[5.](#changes-following-fecal-microbial-transplantation-for-recurrent-cdi) Dynamic changes in short- and long-term bacterial composition following fecal microbiota transplantation for recurrent Clostridium difficile infection -> [dataset link](https://qiita.ucsd.edu/study/description/1924)

[6.](#fecal-transplantation-ibd-and-cdi-dataset) Changes in Microbial Ecology after Fecal Microbiota Transplantation for recurrent C. difficile Infection Affected by Underlying Inflammatory Bowel Disease -> [dataset link](https://qiita.ucsd.edu/study/description/10057)

[7.](#hospital-clinic's-cdi-samples) Dataset of Hospital Clinic's CDI patients' samples pre and post fecal microbiome transplantation 

All studies' data and metadata is available on Qiita platform.

## IBD datasets

Lets upload first study related to Inflammatory bowel disease (Ulcerative colitis and Crohn's disease):

```{r}
# Load metadata
IBD_meta_all <- read.delim(here("00_raw_data/02_IBD", "sample_information_IBD_1.tsv.gz"))  

# Load .tsv.gz files from inflammatory bowel disease analysis
shannon_IBD <- read.delim(here("00_raw_data/02_IBD", "shannon_IBD_1.tsv.gz"))
chao1_IBD <- read.delim(here("00_raw_data/02_IBD", "chao1_IBD_1.tsv.gz"))
fisher_IBD <- read.delim(here("00_raw_data/02_IBD", "fisher_IBD_1.tsv.gz"))
margalef_IBD <- read.delim(here("00_raw_data/02_IBD", "margalef_IBD_1.tsv.gz"))
simpson_IBD <- read.delim(here("00_raw_data/02_IBD", "simpson_IBD_1.tsv.gz"))
gini_IBD <- read.delim(here("00_raw_data/02_IBD", "gini_IBD_1.tsv.gz"))
menhinick_IBD <- read.delim(here("00_raw_data/02_IBD", "menhinick_IBD_1.tsv.gz"))
strong_IBD <- read.delim(here("00_raw_data/02_IBD", "strong_IBD_1.tsv.gz"))
faith_IBD <- read.delim(here("00_raw_data/02_IBD", "faith_IBD_1.tsv.gz"))
pielou_IBD <- read.delim(here("00_raw_data/02_IBD", "pielou_IBD_1.tsv.gz"))
```

```{r}
# Clean and filter metadata
IBD_meta <- dplyr::select(IBD_meta_all, sample_id, sex, race, consent_age, age_at_diagnosis, diagnosis, qiita_study_id)
names(IBD_meta)[names(IBD_meta) == 'diagnosis'] <- 'condition'
names(IBD_meta)[names(IBD_meta) == 'consent_age'] <- 'age'
IBD_meta[IBD_meta == "White"] <-'Caucasian'

IBD_meta <- IBD_meta %>%
  filter(age > 20 & age < 69)

IBD_meta$sample_id <- str_replace(IBD_meta$sample_id, "^.*?\\.", "")


# Merge alpha diversity dataframes
names(faith_IBD)[names(faith_IBD) == 'X.SampleID'] <- 'X'

all_IBD_alpha <- list(shannon_IBD, chao1_IBD, menhinick_IBD, margalef_IBD, fisher_IBD, gini_IBD, strong_IBD, simpson_IBD, faith_IBD, pielou_IBD) 
all_IBD_alpha <- all_IBD_alpha %>% reduce(inner_join, by="X")
names(all_IBD_alpha)[names(all_IBD_alpha) == 'X'] <- 'sample_id'

all_IBD_alpha$sample_id <- str_replace(all_IBD_alpha$sample_id, "^.*?\\.", "")


# Merge alpha metrics' dataframe with metadata
IBD <- all_IBD_alpha %>% left_join(IBD_meta, by=c('sample_id' = 'sample_id'))
IBD <- IBD[complete.cases(IBD),]

nrow(IBD)
table(IBD$condition)
```

After filtering we are left with `r nrow(IBD)` samples for which we have calculated all chosen alpha metrics.

Export data frame into .csv.gz file:
```{r}
write.csv(IBD, gzfile(here("01_tidy_data", "IBD.csv.gz")), row.names=FALSE)
```

### UC samples

Lets upload second study related to Inflammatory bowel disease (Ulcerative colitis):

```{r}
# Load metadata
UC_meta_all <- read.delim(here("00_raw_data/02_IBD", "sample_information_UC_1.tsv.gz"))

# Load .tsv.gz files from inflammatory bowel disease analysis
shannon_UC <- read.delim(here("00_raw_data/02_IBD", "shannon_UC_1.tsv.gz"))
chao1_UC <- read.delim(here("00_raw_data/02_IBD", "chao1_UC_1.tsv.gz"))
fisher_UC <- read.delim(here("00_raw_data/02_IBD", "fisher_UC_1.tsv.gz"))
margalef_UC <- read.delim(here("00_raw_data/02_IBD", "margalef_UC_1.tsv.gz"))
simpson_UC <- read.delim(here("00_raw_data/02_IBD", "simpson_UC_1.tsv.gz"))
gini_UC <- read.delim(here("00_raw_data/02_IBD", "gini_UC_1.tsv.gz"))
menhinick_UC <- read.delim(here("00_raw_data/02_IBD", "menhinick_UC_1.tsv.gz"))
strong_UC <- read.delim(here("00_raw_data/02_IBD", "strong_UC_1.tsv.gz"))
faith_UC <- read.delim(here("00_raw_data/02_IBD", "faith_UC_1.tsv.gz"))
pielou_UC <- read.delim(here("00_raw_data/02_IBD", "pielou_UC_1.tsv.gz"))
```

```{r}
# Clean and filter metadata
UC_meta <- dplyr::select(UC_meta_all, sample_id, sex, race, age, age_diagnosis, disease, qiita_study_id)
names(UC_meta)[names(UC_meta) == 'disease'] <- 'condition'
names(UC_meta)[names(UC_meta) == 'age_diagnosis'] <- 'age_at_diagnosis'
UC_meta[UC_meta == "F"] <- 'female'
UC_meta[UC_meta == "M"] <- 'male'

UC_meta <- UC_meta %>%
  filter(age > 20 & age < 69)

UC_meta$sample_id <- str_replace(UC_meta$sample_id, "^.*?\\.", "")


# Merge alpha diversity dataframes
names(faith_UC)[names(faith_UC) == 'X.SampleID'] <- 'X'

all_UC_alpha <- list(shannon_UC, chao1_UC, menhinick_UC, margalef_UC, fisher_UC, gini_UC, strong_UC, simpson_UC, faith_UC, pielou_UC) 
all_UC_alpha <- all_UC_alpha %>% reduce(inner_join, by="X")
names(all_UC_alpha)[names(all_UC_alpha) == 'X'] <- 'sample_id'

all_UC_alpha$sample_id <- str_replace(all_UC_alpha$sample_id, "^.*?\\.", "")


# Merge alpha metrics' dataframe with metadata
UC <- all_UC_alpha %>% left_join(UC_meta, by=c('sample_id' = 'sample_id'))
UC <- UC[complete.cases(UC),]


nrow(UC)
```

After filtering we are left with `r nrow(UC)` samples for which we have calculated all chosen alpha metrics.

Export data frame into .csv.gz file
```{r}
write.csv(UC, gzfile(here("01_tidy_data", "UC.csv.gz")), row.names=FALSE)
```

## Longitudinal Chron's disease study

Lets upload data from the longitudinal study related to Crohn's disease:

```{r}
# Load metadata
CD_2_all <- read.delim(here("00_raw_data/03_CD_longitudinal", "sample_information_CD_2.tsv.gz"))

# Load .tsv.gz files from inflammatory bowel disease analysis
shannon_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_shannon.tsv.gz"))
chao1_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_chao1.tsv.gz"))
fisher_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_fisher.tsv.gz"))
margalef_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_margalef.tsv.gz"))
simpson_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_simpson.tsv.gz"))
gini_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_gini.tsv.gz"))
menhinick_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_menhinick.tsv.gz"))
strong_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_strong.tsv.gz"))
faith_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_faith.tsv.gz"))
pielou_CD_2 <- read.delim(here("00_raw_data/03_CD_longitudinal", "CD_2_pielou.tsv.gz"))
```

```{r}
# Clean and filter metadata
CD_2_meta <- dplyr::select(CD_2_all, sample_name, sex, age, description, sample_type, weeks, env_package, ibd, surgery_and_ibd, surgery_type, qiita_study_id)

names(CD_2_meta)[names(CD_2_meta) == 'sample_name'] <- 'sample_id'
names(CD_2_meta)[names(CD_2_meta) == 'ibd'] <- 'condition'

CD_2_meta <- CD_2_meta %>%
  filter(age > 20 & age < 69) %>%
  filter(sample_type == "feces")


# Merge alpha diversity data frames
names(faith_CD_2)[names(faith_CD_2) == 'X.SampleID'] <- 'X'

all_CD_2_alpha <- list(shannon_CD_2, chao1_CD_2, fisher_CD_2, margalef_CD_2, gini_CD_2, menhinick_CD_2, strong_CD_2, simpson_CD_2, faith_CD_2, pielou_CD_2)
all_CD_2_alpha <- all_CD_2_alpha %>% reduce(inner_join, by="X")

names(all_CD_2_alpha)[names(all_CD_2_alpha) == 'X'] <- 'sample_id'


# Merge alpha metrics' data frame with metadata
CD_2 <- all_CD_2_alpha %>% left_join(CD_2_meta, by=c('sample_id' = 'sample_id'))
CD_2 <- CD_2[complete.cases(CD_2),]


table(CD_2$description, CD_2$condition)

nrow(CD_2)
```

After filtering we are left with `r nrow(CD_2)` samples for which we have calculated all chosen alpha metrics.

Export data frame into .tsv.gz file:

```{r}
write.csv(CD_2, gzfile(here("01_tidy_data", "CD_2.csv.gz")), row.names=FALSE)
```

## CDI samples 

```{r}
# Load metadata
CDI_meta_all <- read.delim(here("00_raw_data/06_CDI_ncbi", "sample_metadata_ncbi_CDI.tsv.gz"))

# Load .tsv.gz files from inflammatory bowel disease analysis
shannon_ncbi_CDI <- read.delim(here("00_raw_data/06_CDI_ncbi", "shannon_ncbi_CDI.tsv.gz"))
chao1_ncbi_CDI <- read.delim(here("00_raw_data/06_CDI_ncbi", "chao1_ncbi_CDI.tsv.gz"))
fisher_ncbi_CDI <- read.delim(here("00_raw_data/06_CDI_ncbi", "fisher_ncbi_CDI.tsv.gz"))
margalef_ncbi_CDI <- read.delim(here("00_raw_data/06_CDI_ncbi", "margalef_ncbi_CDI.tsv.gz"))
simpson_ncbi_CDI <- read.delim(here("00_raw_data/06_CDI_ncbi", "simpson_ncbi_CDI.tsv.gz"))
gini_ncbi_CDI <- read.delim(here("00_raw_data/06_CDI_ncbi", "gini_ncbi_CDI.tsv.gz"))
menhinick_ncbi_CDI <- read.delim(here("00_raw_data/06_CDI_ncbi", "menhinick_ncbi_CDI.tsv.gz"))
strong_ncbi_CDI <- read.delim(here("00_raw_data/06_CDI_ncbi", "strong_ncbi_CDI.tsv.gz"))
faith_ncbi_CDI <- read.delim(here("00_raw_data/06_CDI_ncbi", "faith_ncbi_CDI.tsv.gz"))
pielou_ncbi_CDI <- read.delim(here("00_raw_data/06_CDI_ncbi", "pielou_ncbi_CDI.tsv.gz"))
```

```{r}
# Clean and filter metadata
CDI_meta <- dplyr::select(CDI_meta_all, sample.id, Age_.years., BMI_.Kg.M2., PPI_use_.0.No_1.Yes.,  Sex_.0.Male_1.Female., Prior_antibiotics_.0.No_1.Yes., Response_to_treatment_.0.success_1.failure., Recurrence_.0.No_1.Yes., SevereCDI_.0.No_1.Yes.)

names(CDI_meta)[names(CDI_meta) == 'sample.id'] <- 'sample_id'
names(CDI_meta)[names(CDI_meta) == 'Age_.years.'] <- 'age'
names(CDI_meta)[names(CDI_meta) == 'BMI_.Kg.M2.'] <- 'BMI'
names(CDI_meta)[names(CDI_meta) == 'PPI_use_.0.No_1.Yes.'] <- 'PPI_use'
names(CDI_meta)[names(CDI_meta) == 'Sex_.0.Male_1.Female.'] <- 'sex'
names(CDI_meta)[names(CDI_meta) == 'Prior_antibiotics_.0.No_1.Yes.'] <- 'prior_antibiotics'
names(CDI_meta)[names(CDI_meta) == 'Response_to_treatment_.0.success_1.failure.'] <- 'response_to_treatment'
names(CDI_meta)[names(CDI_meta) == 'Recurrence_.0.No_1.Yes.'] <- 'recurrence'
names(CDI_meta)[names(CDI_meta) == 'SevereCDI_.0.No_1.Yes.'] <- 'severe_CDI'

CDI_meta <- data.frame(lapply(CDI_meta, as.character), stringsAsFactors=FALSE)
CDI_meta$age <- as.numeric(CDI_meta$age)
CDI_meta$BMI <- as.numeric(CDI_meta$BMI)

CDI_meta <- CDI_meta %>%
    mutate( across(.cols = c(PPI_use, prior_antibiotics, recurrence, severe_CDI),
                  ~str_replace( ., "0", "No"))) %>%
    mutate(across(.cols = c(PPI_use, prior_antibiotics, recurrence, severe_CDI),
                  ~str_replace( ., "1", "Yes")))

CDI_meta <- CDI_meta %>%
    mutate( sex = str_replace( sex, "0", "Male")) %>%
    mutate( sex = str_replace( sex, "1", "Female"))%>%
    mutate( response_to_treatment = str_replace( response_to_treatment, "0", "success")) %>%
    mutate( response_to_treatment = str_replace( response_to_treatment, "1", "failure"))

CDI_meta <- CDI_meta %>%
  filter(age > 20 & age < 69)


# Merge alpha diversity data frames
all_ncbi_CDI_alpha <- list(shannon_ncbi_CDI, chao1_ncbi_CDI, fisher_ncbi_CDI, margalef_ncbi_CDI, gini_ncbi_CDI, menhinick_ncbi_CDI, strong_ncbi_CDI, simpson_ncbi_CDI, faith_ncbi_CDI, pielou_ncbi_CDI) 
all_ncbi_CDI_alpha <- all_ncbi_CDI_alpha %>% reduce(inner_join, by="X")
names(all_ncbi_CDI_alpha)[names(all_ncbi_CDI_alpha) == 'X'] <- 'sample_id'


# Merge alpha metrics' data frame with metadata
ncbi_CDI <- all_ncbi_CDI_alpha %>% left_join(CDI_meta, by=c('sample_id' = 'sample_id'))
ncbi_CDI <- ncbi_CDI[complete.cases(ncbi_CDI),]

ncbi_CDI$condition <- "CDI"

nrow(ncbi_CDI)
```
We have `r nrow(ncbi_CDI)` samples for which we have calculated all chosen alpha metrics.

Export data frame into .tsv.gz file:

```{r}
write.csv(ncbi_CDI, gzfile(here("01_tidy_data", "ncbi_CDI.csv.gz")), row.names=FALSE)
```


# Changes following fecal microbial transplantation for recurrent CDI

Lets load data from study about short- and long-term microbiome changes following FMT for recurrent C. difficile infection:

```{r}
# Load metadata
CDI_FMT_all <- read.delim(here("00_raw_data/04_C_difficile", "sample_information_cdiff.tsv.gz"))

# Load .tsv.gz files from inflammatory bowel disease analysis
shannon_CDI_FMT <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_shannon.tsv.gz"))
chao1_CDI_FMT <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_chao1.tsv.gz"))
fisher_CDI_FMT <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_fisher.tsv.gz"))
margalef_CDI_FMT <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_margalef.tsv.gz"))
simpson_CDI_FMT <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_simpson.tsv.gz"))
gini_CDI_FMT <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_gini.tsv.gz"))
menhinick_CDI_FMT <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_menhinick.tsv.gz"))
strong_CDI_FMT <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_strong.tsv.gz"))
pielou_CDI_FMT <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_pielou.tsv.gz"))
faith_CDI_FMT <- read.delim(here("00_raw_data/04_C_difficile", "cdiff_faith.tsv.gz"))
```

```{r}
# Clean and filter metadata
CDI_FMT_meta <- dplyr::select(CDI_FMT_all, sample_name, animations_subject, sex, race, host_age, host_body_mass_index, day_relative_to_fmt, disease_state, qiita_study_id)

names(CDI_FMT_meta)[names(CDI_FMT_meta) == 'sample_name'] <- 'sample_id'
names(CDI_FMT_meta)[names(CDI_FMT_meta) == 'host_age'] <- 'age'
names(CDI_FMT_meta)[names(CDI_FMT_meta) == 'host_body_mass_index'] <- 'body_mass_index'
names(CDI_FMT_meta)[names(CDI_FMT_meta) == 'day_relative_to_fmt'] <- 'day_since_fmt'

CDI_FMT_meta <- CDI_FMT_meta %>%
  filter(age > 20 & age < 69) %>%
  filter(animations_subject != "Patient" & 
         animations_subject != "Donor")


# Merge alpha diversity data frames
names(faith_CDI_FMT)[names(faith_CDI_FMT) == 'X.SampleID'] <- 'X'

all_CDI_FMT_alpha <- list(shannon_CDI_FMT, chao1_CDI_FMT, fisher_CDI_FMT, margalef_CDI_FMT, gini_CDI_FMT, menhinick_CDI_FMT, strong_CDI_FMT, faith_CDI_FMT, simpson_CDI_FMT, pielou_CDI_FMT)   
all_CDI_FMT_alpha <- all_CDI_FMT_alpha %>% reduce(inner_join, by="X")

names(all_CDI_FMT_alpha)[names(all_CDI_FMT_alpha) == 'X'] <- 'sample_id'


# Merge alpha metrics' data frame with metadata
CDI_FMT <- all_CDI_FMT_alpha %>% left_join(CDI_FMT_meta, by=c('sample_id' = 'sample_id'))
CDI_FMT <- CDI_FMT[complete.cases(CDI_FMT),]


nrow(CDI_FMT)
```

After filtering we are left with `r nrow(CDI_FMT)` samples for which we have calculated all chosen alpha metrics.

```{r}
table(CDI_FMT$body_mass_index, CDI_FMT$animations_subject)
```

Subject CD1 has a BMI of 29.3;
Subject CD2 has a BMI of 27.1;
Subject CD3 has a BMI of 32.5;
Subject CD4 has a BMI of 23.9.

All except CD4 have BMI over 25 which is considered owerveight.
For purpose of comparing improvement after FMT in these four subjects, we are not going to filter out based on BMI.
Furthermore, this is not necessary because we wont compare these samples with healthy dataset.

Export data frame into .tsv.gz file:
```{r}
write.csv(CDI_FMT, gzfile(here("01_tidy_data", "CDI_FMT.csv.gz")), row.names=FALSE)
```

Exported data frames are stored in 01_tidy_data directory of this project. They are going to be used for further statistical analysis.

## Fecal transplantation IBD and CDI dataset

Finally, lets load data from the study about effects of FMT (Fecal Microbiota Transplantation) for recurrent C. difficile infection with underlying inflammatory bowel disease:

```{r}
# Load metadata
FMT_IBD_CDI_all <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "sample_information_tr_CDI_IBD.tsv.gz"))

# Load .tsv.gz files from inflammatory bowel disease analysis
shannon_FMT_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_shannon.tsv.gz"))
chao1_FMT_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_chao1.tsv.gz"))
fisher_FMT_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_fisher.tsv.gz"))
margalef_FMT_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_margalef.tsv.gz"))
simpson_FMT_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_simpson.tsv.gz"))
gini_FMT_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_gini.tsv.gz"))
menhinick_FMT_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_menhinick.tsv.gz"))
strong_FMT_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_strong.tsv.gz"))
faith_FMT_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_faith.tsv.gz"))
pielou_FMT_IBD_CDI <- read.delim(here("00_raw_data/05_FMT_CDI_and_IBD", "tr_IBD_CDI_pielou.tsv.gz"))
```

```{r}
# Clean and filter metadata
FMT_IBD_CDI_meta <- dplyr::select(FMT_IBD_CDI_all, sample_name, animations_subject, sex, age, body_mass_index, pathology, day_since_fmt, donor_or_patient, qiita_study_id)

names(FMT_IBD_CDI_meta)[names(FMT_IBD_CDI_meta) == 'pathology'] <- 'condition'
names(FMT_IBD_CDI_meta)[names(FMT_IBD_CDI_meta) == 'sample_name'] <- 'sample_id'

FMT_IBD_CDI_meta[FMT_IBD_CDI_meta == "Crohn's"] <-'CDI + CD'
FMT_IBD_CDI_meta[FMT_IBD_CDI_meta == "microscopic colitis"] <-'CDI + MC'
FMT_IBD_CDI_meta[FMT_IBD_CDI_meta == "Ulcerative colitis"] <-'CDI + UC'
FMT_IBD_CDI_meta[FMT_IBD_CDI_meta == "no IBD"] <-'CDI'
FMT_IBD_CDI_meta$condition[FMT_IBD_CDI_meta$condition == "Not applicable"] <-'Donors'

FMT_IBD_CDI_meta <- FMT_IBD_CDI_meta %>%
  filter(age > 20 & age < 69) %>%
  filter(body_mass_index > 18.5 &
         body_mass_index < 25)


# Merge alpha diversity data frames
names(faith_FMT_IBD_CDI)[names(faith_FMT_IBD_CDI) == 'X.SampleID'] <- 'X'

all_FMT_IBD_CDI_alpha <- list(shannon_FMT_IBD_CDI, chao1_FMT_IBD_CDI, fisher_FMT_IBD_CDI, margalef_FMT_IBD_CDI, gini_FMT_IBD_CDI, menhinick_FMT_IBD_CDI, strong_FMT_IBD_CDI, simpson_FMT_IBD_CDI, faith_FMT_IBD_CDI, pielou_FMT_IBD_CDI) 
all_FMT_IBD_CDI_alpha <- all_FMT_IBD_CDI_alpha %>% reduce(inner_join, by="X")

names(all_FMT_IBD_CDI_alpha)[names(all_FMT_IBD_CDI_alpha) == 'X'] <- 'sample_id'


# Merge alpha metrics' data frame with metadata
FMT_IBD_CDI <- all_FMT_IBD_CDI_alpha %>% left_join(FMT_IBD_CDI_meta, by=c('sample_id' = 'sample_id'))
FMT_IBD_CDI <- FMT_IBD_CDI[complete.cases(FMT_IBD_CDI),]


nrow(FMT_IBD_CDI)
```

After filtering we are left with `r nrow(FMT_IBD_CDI)` samples for which we have calculated all chosen alpha metrics.

Export data frame into .tsv.gz file:

```{r}
write.csv(FMT_IBD_CDI, gzfile(here("01_tidy_data", "FMT_IBD_CDI.csv.gz")), row.names=FALSE)
```


## Hospital dataset - CDI samples

```{r}
# Load metadata
hosp_CDI_meta_all <- read.delim(here("00_raw_data/07_hospital_CDI", "sample-metadata.tsv.gz"))

# Load .tsv.gz files from inflammatory bowel disease analysis
shannon_hosp_CDI <- read.delim(here("00_raw_data/07_hospital_CDI", "shannon_hosp_CDI.tsv.gz"))
chao1_hosp_CDI <- read.delim(here("00_raw_data/07_hospital_CDI", "chao1_hosp_CDI.tsv.gz"))
fisher_hosp_CDI <- read.delim(here("00_raw_data/07_hospital_CDI", "fisher_hosp_CDI.tsv.gz"))
margalef_hosp_CDI <- read.delim(here("00_raw_data/07_hospital_CDI", "margalef_hosp_CDI.tsv.gz"))
simpson_hosp_CDI <- read.delim(here("00_raw_data/07_hospital_CDI", "simpson_hosp_CDI.tsv.gz"))
gini_hosp_CDI <- read.delim(here("00_raw_data/07_hospital_CDI", "gini_hosp_CDI.tsv.gz"))
menhinick_hosp_CDI <- read.delim(here("00_raw_data/07_hospital_CDI", "menhinick_hosp_CDI.tsv.gz"))
strong_hosp_CDI <- read.delim(here("00_raw_data/07_hospital_CDI", "strong_hosp_CDI.tsv.gz"))
faith_hosp_CDI <- read.delim(here("00_raw_data/07_hospital_CDI", "faith_hosp_CDI.tsv.gz"))
pielou_hosp_CDI <- read.delim(here("00_raw_data/07_hospital_CDI", "pielou_hosp_CDI.tsv.gz"))
```

```{r}
# Clean and filter metadata
hosp_CDI_meta <- dplyr::select(hosp_CDI_meta_all, Sample.ID, sample.purpose, sample.origin, age, time)
names(hosp_CDI_meta)[names(hosp_CDI_meta) == 'Sample.ID'] <- 'sample_id'
names(hosp_CDI_meta)[names(hosp_CDI_meta) == 'sample.origin'] <- 'condition'
names(hosp_CDI_meta)[names(hosp_CDI_meta) == 'time'] <- 'day_since_fmt'

hosp_CDI_meta <- hosp_CDI_meta %>%
  filter(age > 20 & age < 69) %>%
  filter(condition == "Cdif") %>%
  filter(sample.purpose != "Adsorbate")


# Merge alpha diversity data frames
all_hosp_CDI_alpha <- list(shannon_hosp_CDI, chao1_hosp_CDI, fisher_hosp_CDI, margalef_hosp_CDI, gini_hosp_CDI, menhinick_hosp_CDI, strong_hosp_CDI, simpson_hosp_CDI, faith_hosp_CDI, pielou_hosp_CDI) 
all_hosp_CDI_alpha <- all_hosp_CDI_alpha %>% reduce(inner_join, by="X")
names(all_hosp_CDI_alpha)[names(all_hosp_CDI_alpha) == 'X'] <- 'sample_id'

# Merge alpha metrics' data frame with metadata
hosp_CDI <- all_hosp_CDI_alpha %>% left_join(hosp_CDI_meta, by=c('sample_id' = 'sample_id'))
hosp_CDI <- hosp_CDI %>%
  filter(!is.na(sample.purpose))


for(i in 1:nrow(hosp_CDI)) {
  if(hosp_CDI$sample.purpose[i] =="Receptor" & hosp_CDI$day_since_fmt[i] <= 0){
      hosp_CDI$FMT_pre_post[i] <- "pre"
      hosp_CDI$condition[i] <- paste(hosp_CDI$condition[i], "pre", sep="_")
  } else if (hosp_CDI$sample.purpose[i] =="Receptor" & hosp_CDI$day_since_fmt[i] > 0){
      hosp_CDI$FMT_pre_post[i] <- "post"
      hosp_CDI$condition[i] <- paste(hosp_CDI$condition[i], "post", sep="_")
  } else if(hosp_CDI$sample.purpose[i] =="Pool"){
      hosp_CDI$FMT_pre_post[i] <- NA
      hosp_CDI$condition[i] <- "donor"
  }
}

hosp_CDI$FMT_pre_post <- as.factor(hosp_CDI$FMT_pre_post)
hosp_CDI$FMT_pre_post <- relevel(hosp_CDI$FMT_pre_post, "pre")
levels(hosp_CDI$FMT_pre_post)

hosp_CDI <- dplyr::select(hosp_CDI, sample_id, shannon_entropy, chao1, menhinick, margalef, fisher_alpha, simpson, pielou_evenness, gini_index, strong, simpson, faith_pd, condition, FMT_pre_post)

nrow(hosp_CDI)
```
We have `r nrow(hosp_CDI)` samples for which we have calculated all chosen alpha metrics.

Export data frame into .tsv.gz file:

```{r}
write.csv(hosp_CDI, gzfile(here("01_tidy_data", "hosp_CDI.csv.gz")), row.names=FALSE)
```

```{r}
# hosp_CDI_meta_write <- read.delim("~/Desktop/master_project/Data_Hospital_CDI/sample-metadata.tsv")
# 
# 
# hosp_CDI_meta_write$time <- as.numeric(hosp_CDI_meta_write$time)
# 
# for(i in 1:nrow(hosp_CDI_meta_write)) {
#   if(hosp_CDI_meta_write$time[i] <= 0 & !is.na(hosp_CDI_meta_write$time[i])){
#     hosp_CDI_meta_write$FMT.pre.post[i] <- "pre"
#   } else if (hosp_CDI_meta_write$time[i] > 0 & !is.na(hosp_CDI_meta_write$time[i])) {
#     hosp_CDI_meta_write$FMT.pre.post[i] <- "post"
#   }
# }
# 
# write.csv(hosp_CDI_meta_write, gzfile(here("01_tidy_data", "sample-metadata2.tsv")), row.names=FALSE)
```


## Hospital dataset - fecal donor samples

```{r}
# Load metadata
hosp_donor_meta_all <- read.delim(here("00_raw_data/08_hospital_donor", "sample-metadata.tsv.gz"))

# Load .tsv.gz files from inflammatory bowel disease analysis
shannon_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor", "shannon_hosp_donor.tsv.gz"))
chao1_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor", "chao1_hosp_donor.tsv.gz"))
fisher_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor", "fisher_hosp_donor.tsv.gz"))
margalef_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor", "margalef_hosp_donor.tsv.gz"))
simpson_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor", "simpson_hosp_donor.tsv.gz"))
gini_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor", "gini_hosp_donor.tsv.gz"))
menhinick_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor", "menhinick_hosp_donor.tsv.gz"))
strong_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor", "strong_hosp_donor.tsv.gz"))
faith_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor", "faith_hosp_donor.tsv.gz"))
pielou_hosp_donor <- read.delim(here("00_raw_data/08_hospital_donor", "pielou_hosp_donor.tsv.gz"))
```

```{r}
# Clean and filter metadata
hosp_donor_meta <- dplyr::select(hosp_donor_meta_all, SampleID, Edad, Genero, n, Donor, n_donante)
names(hosp_donor_meta)[names(hosp_donor_meta) == 'SampleID'] <- 'sample_id'
names(hosp_donor_meta)[names(hosp_donor_meta) == 'Edad'] <- 'age'
names(hosp_donor_meta)[names(hosp_donor_meta) == 'Genero'] <- 'sex'
names(hosp_donor_meta)[names(hosp_donor_meta) == 'Donor'] <- 'donor'
names(hosp_donor_meta)[names(hosp_donor_meta) == 'n_donante'] <- 'n_donation'

# Merge alpha diversity data frames
all_hosp_donor_alpha <- list(shannon_hosp_donor, chao1_hosp_donor, fisher_hosp_donor, margalef_hosp_donor, gini_hosp_donor, menhinick_hosp_donor, strong_hosp_donor, simpson_hosp_donor, faith_hosp_donor, pielou_hosp_donor) 
all_hosp_donor_alpha <- all_hosp_donor_alpha %>% reduce(inner_join, by="X")
names(all_hosp_donor_alpha)[names(all_hosp_donor_alpha) == 'X'] <- 'sample_id'

# Merge alpha metrics' data frame with metadata
hosp_donor <- all_hosp_donor_alpha %>% left_join(hosp_donor_meta, by=c('sample_id' = 'sample_id'))
hosp_donor <- hosp_donor[complete.cases(hosp_donor),]

hosp_donor$condition <- "healthy_donors"
nrow(hosp_donor)
```

Export table:
```{r}
write.csv(hosp_donor, gzfile(here("01_tidy_data", "hosp_donor.csv.gz")), row.names=FALSE)
```


# Session information

```{r}
sessionInfo()
```

# References

