---
title: "Hospital_Clinic's_data"
author: "Natasa Mortvanski"
date: "2023-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Compare Hospital Clinic's data with other datasets

Loading libraries:

```{r}
#install.packages("readr")
library("readr")
#install.packages('data.table')
library(data.table)
library(stringr)
#install.packages('ggplot2')
library(ggplot2)
library(gridExtra)
library(grid)
library(plyr)
library(purrr)
library(dplyr)
#install.packages('flextable')
library(flextable)
library(tibble)
library(ComplexHeatmap)
library(RColorBrewer)
library(ggplotify)
library(here)
library(table1)
library(tidyverse)
```

Load data: 

```{r}
# Load healthy samples' table
all_healthy <- read.csv(here("01_tidy_data", "AGP_healthy.csv.gz"), header = TRUE, sep = ",")

all_healthy <- dplyr::select(all_healthy, sample_id, shannon_entropy, chao1, menhinick, margalef, fisher_alpha, simpson, pielou_evenness, gini_index, strong, simpson, faith_pd, sex, race, host_age, condition)

names(all_healthy)[names(all_healthy) == 'host_age'] <- 'age'

#Load *CDI data from NCBI* data
CDI <- read.csv(here("01_tidy_data", "ncbi_CDI.csv.gz"), header = TRUE, sep = ",")

# Load CDI and donor data from Hospital Clínic
hospital_CDI <- read.csv(here("01_tidy_data", "hosp_CDI.csv.gz"), header = TRUE, sep = ",")
hospital_donor <- read.csv(here("01_tidy_data", "hosp_donor.csv.gz"), header = TRUE, sep = ",")
```


Let's define vector of names of the alpha diversity metrics that are going to be analysed:

```{r}
metric <- c("chao1", "margalef", "menhinick", "fisher_alpha", "faith_pd", "gini_index", "strong", "pielou_evenness", "shannon_entropy", "simpson") 
```

Let's define function for plotting violin plots that we are going to use in the whole analysis:

```{r}
plot_violin <- function(df, column){
  violin <- vector('list', length(metric))
  
  for (i in 1:length(metric)){
    mean_line <- df %>% dplyr::group_by(.data[[column]]) %>% dplyr::summarise(grp_mean=mean(.data[[metric[i]]]))
    plot_data <- df %>% dplyr::group_by(.data[[column]]) %>% dplyr::mutate(m = mean(.data[[metric[i]]])) 
  
    violin[[i]] <- plot_data %>% ggplot(aes(x = .data[[metric[i]]], y = reorder(.data[[column]], -m), color = .data[[column]], fill = .data[[column]])) +
      geom_violin()+
      geom_boxplot(width=0.1, color="grey", alpha=0.2) +
      geom_vline(data = mean_line, aes(xintercept = grp_mean, color = .data[[column]]), linetype = "dashed")+ 
      labs(x = metric[i])+
      ylab("")+
      theme(legend.position="none") 
    
    if(metric[i] != "shannon_entropy" & metric[i] != "strong" & metric[i] != "gini_index"  & metric[i] != "menhinick"){
     violin[[i]] <- violin[[i]] + 
       scale_x_continuous(trans = 'log10') +
       xlab(paste(metric[i], "(log10)", sep = " ")) 
    }
  }
  return(violin)
}
```

Function for doing Mann-Whitney-Wilcoxon test:

```{r}
do_wilcox_test <- function(df, column){
  test <- list()
  
  for (i in 1:length(metric)){
    test[[i]] <- pairwise.wilcox.test(df[[metric[i]]], df[[column]], p.adjust.method="none") %>% 
    broom::tidy() %>% add_column(parameter = metric[i], .before='group1')
    test[[i]]$p.value <- round(test[[i]]$p.value, digits = 5)
  }
  
  tests <- do.call(what = rbind, args = test)
  
  return(tests)
}
```

First, lets examine difference in alpha diversity before and after FMT in CDI data set:

```{r}
table1(~ chao1 + margalef + menhinick + fisher_alpha + faith_pd + gini_index + strong + pielou_evenness + shannon_entropy + simpson | condition, data=hospital_CDI)
```

```{r}
violin_hospital_2 <- vector('list', length(metric))

violin_hospital_2 <- plot_violin(hospital_CDI, "condition")

grid.arrange(violin_hospital_2[[1]], violin_hospital_2[[2]], violin_hospital_2[[3]], violin_hospital_2[[4]], violin_hospital_2[[5]], violin_hospital_2[[6]], violin_hospital_2[[7]], violin_hospital_2[[8]], violin_hospital_2[[9]], violin_hospital_2[[10]], ncol=3) 
```

```{r}
test_CDI_hospital <- list()

test_CDI_hospital <- do_wilcox_test(hospital_CDI, "condition")

test_CDI_hospital_1 <- test_CDI_hospital %>% 
  add_column(p.adjusted = round(p.adjust(test_CDI_hospital$p.value, "fdr"), digits=5), .after='p.value') %>%
  arrange(p.value, parameter)  %>%
  flextable() %>% 
  bold(~ p.value < 0.05, 4) %>%
  bold(~ p.adjusted < 0.05, 5) %>%
  add_header_lines(values = "Results of the Mann-Whitney-Wilcoxon test for distributions of Hospital Clínic's CDI samples vs healthy control")

test_CDI_hospital_1
```

Lets see how different is alpha diversity of first and second dataset's donor data from Hospital Clínic:

```{r}
hospital_donor_first <- hospital_CDI %>%
  filter(condition == "donor")

compare_donors <- rbind.fill(hospital_donor, hospital_donor_first)

compare_donors$condition <- as.factor(compare_donors$condition)

# Sizes of each dataset
table(compare_donors$condition)
```

```{r}
for (i in 1:length(metric)){
p<- compare_donors %>% ggplot(aes(x=.data[[metric[i]]], color=condition)) +
  geom_density()

print(p)
}
```

```{r}
test_donors <- list()

test_donors <- do_wilcox_test(compare_donors, "condition")

test_donors_1 <- test_donors %>% 
  add_column(p.adjusted = round(p.adjust(test_donors$p.value, "fdr"), digits=5), .after='p.value') %>%
  arrange(p.value, parameter)  %>%
  flextable() %>% 
  bold(~ p.value < 0.05, 4) %>%
  bold(~ p.adjusted < 0.05, 5) %>%
  add_header_lines(values = "Results of the Mann-Whitney-Wilcoxon test for distributions of Hospital Clínic's CDI samples vs healthy control")

test_donors_1
```

Lets see how alpha diversity of CDI data before FMT diverges from stool donor data from Hospital Clínic:

```{r}
hospital_CDI_pre_FMT <- hospital_CDI %>%
  filter(FMT_pre_post == "pre")

compare_hospital <- rbind.fill(hospital_donor, hospital_CDI_pre_FMT)

compare_hospital$condition <- as.factor(compare_hospital$condition)

# Sizes of each dataset
table(compare_hospital$condition)
```

```{r}
table1(~ chao1 + margalef + menhinick + fisher_alpha + faith_pd + gini_index + strong + pielou_evenness + shannon_entropy + simpson | condition, data=compare_hospital)
```

```{r}
violin_compare <- vector('list', length(metric))

violin_compare <- plot_violin(compare_hospital, "condition")

#violin_compare

grid.arrange(violin_compare[[1]], violin_compare[[2]], violin_compare[[3]], violin_compare[[4]], violin_compare[[5]], violin_compare[[6]], violin_compare[[7]], violin_compare[[8]], violin_compare[[9]], violin_compare[[10]], ncol=3) 
```

```{r}
test_CDI_healthy <- list()

test_CDI_healthy <- do_wilcox_test(compare_hospital, "condition")

test_CDI_healthy_1 <- test_CDI_healthy %>% 
  add_column(p.adjusted = round(p.adjust(test_CDI_healthy$p.value, "fdr"), digits=5), .after='p.value') %>%
  arrange(p.value, parameter)  %>%
  flextable() %>% 
  bold(~ p.value < 0.05, 4) %>%
  bold(~ p.adjusted < 0.05, 5) %>%
  add_header_lines(values = "Results of the Mann-Whitney-Wilcoxon test for distributions of Hospital Clínic's CDI samples vs healthy control")

test_CDI_healthy_1
```


Compare Hospital's data to AGP healthy data and CDI dataset from BioProject:

```{r}
all_data_comparison <- rbind.fill(all_healthy, CDI, hospital_donor, hospital_CDI)
  
all_data_comparison <- dplyr::select(all_data_comparison, shannon_entropy, chao1, menhinick, margalef, fisher_alpha, simpson, gini_index, strong, pielou_evenness, faith_pd, condition)

all_data_comparison$condition[all_data_comparison$condition=="healthy"] <- "healthy_AGP"
all_data_comparison$condition[all_data_comparison$condition=="CDI"] <- "CDI_BioProject"
all_data_comparison$condition[all_data_comparison$condition=="Cdif_post"] <- "hospital_CDI_post"
all_data_comparison$condition[all_data_comparison$condition=="Cdif_pre"] <- "hospital_CDI_pre"

violin_hospital_compare <- vector('list', length(metric))

violin_hospital_compare <- plot_violin(all_data_comparison, "condition")


grid.arrange(violin_hospital_compare[[1]], violin_hospital_compare[[2]], violin_hospital_compare[[3]], violin_hospital_compare[[4]], ncol=2) 
grid.arrange(violin_hospital_compare[[5]], violin_hospital_compare[[6]], violin_hospital_compare[[7]], violin_hospital_compare[[8]], ncol=2) 
grid.arrange(violin_hospital_compare[[9]], violin_hospital_compare[[10]], ncol=2)

table(all_data_comparison$condition)
```
