---
title: "Hospital_Clinic's_data"
author: "Natasa Mortvanski"
date: "2023-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Compare Hospital Clinic's data with other datasets

Loading libraries:

```{r}
#install.packages("readr")
library("readr")
#install.packages('data.table')
library(data.table)
library(stringr)
#install.packages('ggplot2')
library(ggplot2)
library(gridExtra)
library(grid)
library(plyr)
library(purrr)
library(dplyr)
#install.packages('flextable')
library(flextable)
library(tibble)
library(ComplexHeatmap)
library(RColorBrewer)
library(ggplotify)
library(here)
library(table1)
library(tidyverse)
```

Load data: 

```{r}
# Load healthy samples' table
all_healthy <- read.csv(here("01_tidy_data", "AGP_healthy.csv.gz"), header = TRUE, sep = ",")

all_healthy <- dplyr::select(all_healthy, sample_id, shannon_entropy, chao1, menhinick, margalef, fisher_alpha, simpson, pielou_evenness, gini_index, strong, simpson, faith_pd, sex, race, host_age, condition)

names(all_healthy)[names(all_healthy) == 'host_age'] <- 'age'

#Load *CDI data from NCBI* data
CDI <- read.csv(here("01_tidy_data", "ncbi_CDI.csv.gz"), header = TRUE, sep = ",")

# Load CDI and donor data from Hospital Clínic
hospital_CDI <- read.csv(here("01_tidy_data", "hosp_CDI.csv.gz"), header = TRUE, sep = ",")
hospital_donor <- read.csv(here("01_tidy_data", "hosp_donor.csv.gz"), header = TRUE, sep = ",")

hospital_CDI_donors <- hospital_CDI %>%
  filter(condition == "donor")

hospital_CDI <- rbind.fill(hospital_donor, hospital_CDI)
hospital_CDI$condition[hospital_CDI$condition=="Cdif_post"] <- "CDIpost"
hospital_CDI$condition[hospital_CDI$condition=="Cdif_pre"] <- "CDIpre"
hospital_CDI$condition[hospital_CDI$condition=="healthy_donors"] <- "donor"

```


Let's define vector of names of the alpha diversity metrics that are going to be analysed:

```{r}
metric <- c("chao1", "margalef", "menhinick", "fisher_alpha", "faith_pd", "gini_index", "strong", "pielou_evenness", "shannon_entropy", "simpson") 
```

Let's define function for plotting violin plots that we are going to use in the whole analysis:

```{r}
plot_violin <- function(df, column){
  violin <- vector('list', length(metric))
  
  for (i in 1:length(metric)){
    mean_line <- df %>% dplyr::group_by(.data[[column]]) %>% dplyr::summarise(grp_mean=mean(.data[[metric[i]]]))
    plot_data <- df %>% dplyr::group_by(.data[[column]]) %>% dplyr::mutate(m = mean(.data[[metric[i]]])) 
  
    violin[[i]] <- plot_data %>% ggplot(aes(x = .data[[metric[i]]], y = reorder(.data[[column]], -m), color = .data[[column]], fill = .data[[column]])) +
      geom_violin()+
      geom_boxplot(width=0.1, color="grey", alpha=0.2) +
      geom_vline(data = mean_line, aes(xintercept = grp_mean, color = .data[[column]]), linetype = "dashed")+ 
      labs(x = metric[i])+
      ylab("")+
      theme(legend.position="none") 
    
    if(metric[i] != "shannon_entropy" & metric[i] != "strong" & metric[i] != "gini_index"  & metric[i] != "menhinick"){
     violin[[i]] <- violin[[i]] + 
       scale_x_continuous(trans = 'log10') +
       xlab(paste(metric[i], "(log10)", sep = " ")) 
    }
  }
  return(violin)
}
```

Function for doing Mann-Whitney-Wilcoxon test:

```{r}
do_wilcox_test <- function(df, column){
  test <- list()
  
  for (i in 1:length(metric)){
    test[[i]] <- pairwise.wilcox.test(df[[metric[i]]], df[[column]], p.adjust.method="none") %>% 
    broom::tidy() %>% add_column(parameter = metric[i], .before='group1')
    test[[i]]$p.value <- round(test[[i]]$p.value, digits = 5)
  }
  
  tests <- do.call(what = rbind, args = test)
  
  return(tests)
}
```

First, lets examine difference in alpha diversity before and after FMT in CDI data set:

```{r}
table1(~ chao1 + margalef + menhinick + fisher_alpha + faith_pd + gini_index + strong + pielou_evenness + shannon_entropy + simpson | condition, data=hospital_CDI)
```

```{r}
violin_hospital_2 <- vector('list', length(metric))

violin_hospital_2 <- plot_violin(hospital_CDI, "condition")

grid.arrange(violin_hospital_2[[1]], violin_hospital_2[[2]], violin_hospital_2[[3]], violin_hospital_2[[4]], violin_hospital_2[[5]], violin_hospital_2[[6]], violin_hospital_2[[7]], violin_hospital_2[[8]], violin_hospital_2[[9]], violin_hospital_2[[10]], ncol=3) 
```

```{r}
library(ggpubr)

my_comparisons <- list( c("CDIpost","CDIpre"),  c("CDIpost","donor"), c("CDIpre","donor"))

violin <- vector('list', length(metric))

for (i in 1:length(metric)){

  violin[[i]] <- ggviolin(hospital_CDI, x = "condition", y = metric[i], fill = "condition", alpha = 0.8,
           palette = "Dark2", order = c("CDIpre", "CDIpost", "donor"),
           add = c("boxplot"), add.params = list(fill = "white")) +
    stat_compare_means(comparisons = my_comparisons, label = "p.signif") +
    #stat_compare_means (label.y = max(hospital_CDI[[metric[i]]]*1.1)) +
    xlab("")+
    font("xy.text", size = 10) +
    rremove("legend") 
}

grid.arrange(violin[[1]], violin[[2]], violin[[3]], violin[[4]], violin[[5]], violin[[6]], ncol=3) 
grid.arrange(violin[[7]], violin[[8]], violin[[9]], violin[[10]], ncol=3) 

#table(hospital_CDI$condition)
```

```{r}
box <- vector('list', length(metric))

for (i in 1:length(metric)){

  box[[i]] <- ggboxplot(hospital_CDI, x = "condition", y = metric[i], color = "condition",
           palette = "Dark2", order = c("CDIpre", "CDIpost", "donor")) +
    stat_compare_means(comparisons = my_comparisons, label = "p.signif") +
    xlab("")+
    font("xy.text", size = 10) +
    rremove("legend") 
}

grid.arrange(box[[1]], box[[2]], box[[3]], box[[4]], box[[5]], box[[6]], ncol=3) 
grid.arrange(box[[7]], box[[8]], box[[9]], box[[10]], ncol=3) 
```

```{r}
test_CDI_hospital <- list()

test_CDI_hospital <- do_wilcox_test(hospital_CDI, "condition")

test_CDI_hospital_1 <- test_CDI_hospital %>% 
  add_column(p.adjusted = round(p.adjust(test_CDI_hospital$p.value, "fdr"), digits=5), .after='p.value') %>%
  arrange(p.value, parameter)  %>%
  flextable() %>% 
  bold(~ p.value < 0.05, 4) %>%
  bold(~ p.adjusted < 0.05, 5) %>%
  add_header_lines(values = "Results of the Mann-Whitney-Wilcoxon test for distributions of Hospital Clínic's CDI samples vs healthy control")

test_CDI_hospital_1
```

Lets see how different is alpha diversity of first and second dataset's donor data from Hospital Clínic:

```{r}
compare_donors <- rbind.fill(hospital_donor, hospital_CDI)

compare_donors$condition <- as.factor(compare_donors$condition)

# Sizes of each dataset
table(compare_donors$condition)
```

```{r}
box_sig <- vector('list', length(metric))
my_comparisons <- list(c("healthy_donors","donor"), c("healthy_donors","CDIpost"), c("healthy_donors","CDIpre"), c("donor","CDIpost"), c("donor","CDIpre"), c("CDIpost","CDIpre"))

for (i in 1:length(metric)){
  
box_sig[[i]] <- ggboxplot(compare_donors, x = "condition", y = metric[i], color = "condition",
           palette = "Dark2", order = c("CDIpre", "CDIpost", "healthy_donors", "donor"),
           outlier.shape =NA) +
    stat_compare_means(comparisons = my_comparisons, label = "p.signif") +
    xlab("")+
    font("xy.text", size = 7) +
    rremove("legend") 
}

grid.arrange(box_sig[[1]], box_sig[[2]], box_sig[[3]], box_sig[[4]], box_sig[[5]], ncol=3) 
grid.arrange(box_sig[[6]], box_sig[[7]], box_sig[[8]], box_sig[[9]], box_sig[[10]], ncol=3) 

```


```{r}
test_donors <- list()

test_donors <- do_wilcox_test(compare_donors, "condition")

test_donors_1 <- test_donors %>% 
  add_column(p.adjusted = round(p.adjust(test_donors$p.value, "fdr"), digits=5), .after='p.value') %>%
  arrange(p.value, parameter)  %>%
  flextable() %>% 
  bold(~ p.value < 0.05, 4) %>%
  bold(~ p.adjusted < 0.05, 5) %>%
  add_header_lines(values = "Results of the Mann-Whitney-Wilcoxon test for distributions of Hospital Clínic's CDI samples vs healthy control")

test_donors_1
```

Compare Hospital's data to AGP healthy data and CDI dataset from BioProject:

```{r}
all_data_comparison <- rbind.fill(all_healthy, CDI, hospital_CDI)
  
all_data_comparison <- dplyr::select(all_data_comparison, shannon_entropy, chao1, menhinick, margalef, fisher_alpha, simpson, gini_index, strong, pielou_evenness, faith_pd, condition)

all_data_comparison$condition[all_data_comparison$condition=="healthy"] <- "AGP"

violin_hospital_compare <- vector('list', length(metric))

violin_hospital_compare <- plot_violin(all_data_comparison, "condition")

# PNG device
#jpeg("my_plot.jpeg", quality = 100)

grid.arrange(violin_hospital_compare[[1]], violin_hospital_compare[[2]], violin_hospital_compare[[3]], violin_hospital_compare[[4]], ncol=2) 

#dev.off()

grid.arrange(violin_hospital_compare[[5]], violin_hospital_compare[[6]], violin_hospital_compare[[7]], violin_hospital_compare[[8]], ncol=2) 
grid.arrange(violin_hospital_compare[[9]], violin_hospital_compare[[10]], ncol=2)

grid.arrange(violin_hospital_compare[[1]], violin_hospital_compare[[2]], violin_hospital_compare[[3]], violin_hospital_compare[[4]], violin_hospital_compare[[5]], violin_hospital_compare[[6]], violin_hospital_compare[[7]], violin_hospital_compare[[8]], violin_hospital_compare[[9]], ncol=3) 

table(all_data_comparison$condition)
```
```{r}
library(ggpubr)
my_comparisons <- list( c("AGP","CDI"),  c("AGP","donor"), c("AGP","CDIpost"), c("AGP","CDIpre"),  c("CDI","donor"), c("CDI","CDIpost"), c("CDI","CDIpre"), c("donor","CDIpost"), c("donor","CDIpre"), c("CDIpost","CDIpre"))

ggviolin(all_data_comparison, x = "condition", y = "simpson", fill = "condition",
         palette = "Dark2",
         add = "boxplot", add.params = list(fill = "white"))+
  stat_compare_means(comparisons = my_comparisons, label = "p.signif") # Add significance levels
  #stat_compare_means(label.y = 50)                                      # Add global the p-value 

combn(unique(all_data_comparison$condition), 2, FUN=paste, collapse=',', simplify=FALSE)
```

```{r}
box_sig <- vector('list', length(metric))
my_comparisons <- list( c("AGP","CDI"),  c("AGP","donor"), c("AGP","CDIpost"), c("AGP","CDIpre"),  c("CDI","donor"), c("CDI","CDIpost"), c("CDI","CDIpre"), c("donor","CDIpost"), c("donor","CDIpre"), c("CDIpost","CDIpre"))

for (i in 1:length(metric)){
  
box_sig[[i]] <- ggboxplot(all_data_comparison, x = "condition", y = metric[i], color = "condition",
           palette = "Dark2", order = c("CDI", "CDIpre", "CDIpost", "donor", "AGP"),
           outlier.shape =NA) +
    stat_compare_means(comparisons = my_comparisons, label = "p.signif") +
    xlab("")+
    font("xy.text", size = 7) +
    rremove("legend") 
}

grid.arrange(box_sig[[1]], box_sig[[2]], box_sig[[3]], box_sig[[4]], box_sig[[5]], ncol=3) 
grid.arrange(box_sig[[6]], box_sig[[7]], box_sig[[8]], box_sig[[9]], box_sig[[10]], ncol=3) 

```

```{r}
box_2 <- vector('list', length(metric))
  
for (i in 1:length(metric)){
  plot_data <- all_data_comparison %>% dplyr::group_by(condition) %>% dplyr::mutate(m = mean(.data[[metric[i]]])) 

  box_2[[i]] <- plot_data %>% ggplot(aes( x = reorder(condition, m), y = .data[[metric[i]]], color = condition)) +
    geom_boxplot()+ #outlier.shape = NA
    labs(x = metric[i])+
    xlab("")+
    theme_classic()+
    font("xy.text", size = 7) +
    #rotate_x_text(30) +
    rremove("legend") +
    scale_fill_brewer(palette="Dark2") 
}

grid.arrange(box_2[[1]], box_2[[2]], box_2[[3]], box_2[[4]], box_2[[5]], box_2[[6]], ncol=3) 
grid.arrange(box_2[[7]], box_2[[8]], box_2[[9]], box_2[[10]], ncol=3) 

```
