---
title: "Hospital_Clinic's_data"
author: "Natasa Mortvanski"
date: "2023-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Compare Hospital Clinic's data with other datasets

Loading libraries:

```{r}
#install.packages("readr")
library("readr")
#install.packages('data.table')
library(data.table)
library(stringr)
#install.packages('ggplot2')
library(ggplot2)
library(gridExtra)
library(grid)
library(plyr)
library(purrr)
library(dplyr)
#install.packages('flextable')
library(flextable)
library(tibble)
library(ComplexHeatmap)
library(RColorBrewer)
library(ggplotify)
library(here)
library(table1)
library(tidyverse)
library(writexl)
```

Load data: 

```{r}
# Load healthy samples' table
all_healthy <- read.csv(here("01_tidy_data", "AGP_healthy.csv.gz"), header = TRUE, sep = ",")

all_healthy <- dplyr::select(all_healthy, sample_id, shannon_entropy, chao1, menhinick, margalef, fisher_alpha, simpson, pielou_evenness, gini_index, strong, simpson, faith_pd, sex, race, host_age, condition)

names(all_healthy)[names(all_healthy) == 'host_age'] <- 'age'

#Load *CDI data from NCBI* data
CDI <- read.csv(here("01_tidy_data", "ncbi_CDI.csv.gz"), header = TRUE, sep = ",")

# Load CDI and donor data from Hospital Clínic
hospital_CDI <- read.csv(here("01_tidy_data", "hosp_CDI.csv.gz"), header = TRUE, sep = ",")
hospital_donor <- read.csv(here("01_tidy_data", "hosp_donor.csv.gz"), header = TRUE, sep = ",")

hospital_CDI_donors <- hospital_CDI %>%
  filter(condition == "donor")

hospital_CDI <- rbind.fill(hospital_donor, hospital_CDI)
hospital_CDI$condition[hospital_CDI$condition=="Cdif_post"] <- "CDIpost"
hospital_CDI$condition[hospital_CDI$condition=="Cdif_pre"] <- "CDIpre"
hospital_CDI$condition[hospital_CDI$condition=="healthy_donors"] <- "donor"

```

Export table:
```{r}
write.csv(hospital_CDI, gzfile(here("01_tidy_data", "hosp_donor_and_CDI.csv.gz")), row.names=FALSE)
```

Let's define vector of names of the alpha diversity metrics that are going to be analysed:

```{r}
metric <- c("chao1", "margalef", "menhinick", "fisher_alpha", "faith_pd", "gini_index", "strong", "pielou_evenness", "shannon_entropy", "simpson") 
```

Let's define function for plotting violin plots that we are going to use in the whole analysis:

```{r}
plot_violin <- function(df, column){
  violin <- vector('list', length(metric))
  
  for (i in 1:length(metric)){
    mean_line <- df %>% dplyr::group_by(.data[[column]]) %>% dplyr::summarise(grp_mean=mean(.data[[metric[i]]]))
    plot_data <- df %>% dplyr::group_by(.data[[column]]) %>% dplyr::mutate(m = mean(.data[[metric[i]]])) 
  
    violin[[i]] <- plot_data %>% ggplot(aes(x = .data[[metric[i]]], y = reorder(.data[[column]], -m), color = .data[[column]], fill = .data[[column]])) +
      geom_violin()+
      geom_boxplot(width=0.1, color="grey", alpha=0.2) +
      geom_vline(data = mean_line, aes(xintercept = grp_mean, color = .data[[column]]), linetype = "dashed")+ 
      xlab(if (metric[i] == "shannon_entropy"){"Shannon entropy (❋)"} else if (metric[i] =="chao1"){"Chao1 (+)"} else if (metric[i] == "menhinick"){"Menhinick (+)"} else if (metric[i] == "margalef"){"Margalef (+)"} else if (metric[i] == "fisher_alpha"){"Fisher alpha (+)"} else if (metric[i] == "simpson"){"Simpson (❋)"} else if (metric[i] == "gini_index"){"Gini index (x)"} else if (metric[i] == "strong"){"Strong dominance (x)"} else if (metric[i] == "pielou_evenness"){"Pielou evenness (x)"} else if (metric[i] == "faith_pd"){"Faith PD (+)"}) +
      ylab("")+
      theme(legend.position="none") 
    
  }
  return(violin)
}
```

```{r}
label_fun <- function(x){
  if (x == "shannon_entropy"){y <- "Shannon entropy (❋)"} else if (x =="chao1"){y <- "Chao1 (+)"} else if (x == "menhinick"){y <- "Menhinick (+)"} else if (x == "margalef"){y <- "Margalef (+)"} else if (x == "fisher_alpha"){y <- "Fisher alpha (+)"} else if (x == "simpson"){y <- "Simpson (❋)"} else if (x == "gini_index"){y <- "Gini index (x)"} else if (x == "strong"){y <- "Strong dominance (x)"} else if (x == "pielou_evenness"){y <- "Pielou evenness (x)"} else if (x == "faith_pd"){y <- "Faith PD (+)"}
  return(y)
}
```

Function for doing Mann-Whitney-Wilcoxon test:

```{r}
do_wilcox_test <- function(df, column){
  test <- list()
  
  for (i in 1:length(metric)){
    test[[i]] <- pairwise.wilcox.test(df[[metric[i]]], df[[column]], p.adjust.method="none") %>% 
    broom::tidy() %>% add_column(parameter = metric[i], .before='group1')
    #test[[i]]$p.value <- round(test[[i]]$p.value, digits = 16)
  }
  
  tests <- do.call(what = rbind, args = test)
  
  return(tests)
}
```

First, lets examine difference in alpha diversity before and after FMT in CDI data set:

```{r}
table1(~ chao1 + margalef + menhinick + fisher_alpha + faith_pd + gini_index + strong + pielou_evenness + shannon_entropy + simpson | condition, data=hospital_CDI)
```

```{r}
violin_hospital_2 <- vector('list', length(metric))

violin_hospital_2 <- plot_violin(hospital_CDI, "condition")

grid.arrange(violin_hospital_2[[1]], violin_hospital_2[[2]], violin_hospital_2[[3]], violin_hospital_2[[4]], violin_hospital_2[[5]], violin_hospital_2[[6]], violin_hospital_2[[7]], violin_hospital_2[[8]], violin_hospital_2[[9]], violin_hospital_2[[10]], ncol=3) 
```

```{r}
library(ggpubr)

my_comparisons <- list( c("CDIpost","CDIpre"),  c("CDIpost","donor"))

violin_1 <- vector('list', length(metric))

for (i in 1:length(metric)){

  violin_1[[i]] <- ggviolin(hospital_CDI, x = "condition", y = metric[i], color = "condition", fill = "condition", 
           palette = "Dark2", order = c("CDIpre", "CDIpost", "donor"),
           add = c("boxplot"), add.params = list(color = "black", fill = "white", width=0.10,  alpha = 0.8)) +
    stat_compare_means(comparisons = my_comparisons, label = "p.signif",  vjust = 1.5) +
    xlab("")+
    ylab(label_fun(metric[i])) +
    font("xy.text", size = 10) +
    rremove("legend") +
    theme(axis.title.y=element_text(size=10))
}

grid.arrange(violin_1[[1]], violin_1[[2]], violin_1[[3]], violin_1[[4]], violin_1[[5]], violin_1[[6]], ncol=3) 
grid.arrange(violin_1[[7]], violin_1[[8]], violin_1[[9]], violin_1[[10]], ncol=3) 

```
```{r}
#install.packages("patchwork")
# library(patchwork)
# 
# violin[[1]] + violin[[2]] + violin[[3]] + violin[[4]] + violin[[5]] + violin[[6]] 
# patch2 <- violin[[7]] + violin[[8]] + violin[[9]] + violin[[10]]

```

```{r}
box_1 <- vector('list', length(metric))

for (i in 1:length(metric)){

  box_1[[i]] <- ggboxplot(hospital_CDI, x = "condition", y = metric[i], color = "condition", order = c("CDIpre", "CDIpost", "donor")) +
    stat_compare_means(comparisons = my_comparisons, label = "p.signif",  vjust = 1.5) +
    xlab("")+
    ylab(label_fun(metric[i])) +
    font("xy.text", size = 10) +
    rremove("legend") +
    scale_color_manual(values=c('#a6611a','#80cdc1','#018571'))
}

#png(file = here("03_plots_and_tables", "box_plot_1_hosp.png"), width = 3000, height = 2000, res = 300)
grid.arrange(box_1[[1]], box_1[[2]], box_1[[3]], box_1[[4]], box_1[[5]], box_1[[6]], ncol=3) 
#dev.off()

#png(file = here("03_plots_and_tables", "box_plot_2_hosp.png"), width = 3000, height = 2000, res = 300)
grid.arrange( box_1[[7]], box_1[[8]], box_1[[9]], box_1[[10]], ncol=3) 
#dev.off()
```

```{r}
test_CDI_hospital <- list()

test_CDI_hospital <- do_wilcox_test(hospital_CDI, "condition")

test_CDI_hospital_1 <- test_CDI_hospital %>% 
  #add_column(p.adjusted = round(p.adjust(test_CDI_hospital$p.value, "fdr"), digits=16), .after='p.value') %>%
  add_column(p.adjusted = p.adjust(test_CDI_hospital$p.value, "fdr"), .after='p.value') %>%
  arrange(p.value, parameter)  

test_CDI_hospital_1%>%
  flextable() %>% 
  bold(~ p.value < 0.05, 4) %>%
  bold(~ p.adjusted < 0.05, 5) %>%
  add_header_lines(values = "Results of the Mann-Whitney-Wilcoxon test for distributions of Hospital Clínic's CDI samples vs healthy control")

write_xlsx(test_CDI_hospital_1, here("03_plots_and_tables", "CDI_hospital.xlsx"))

```


Compare Hospital's data to AGP healthy data and CDI dataset from BioProject:

```{r}
all_data_comparison <- rbind.fill(all_healthy, CDI, hospital_CDI)
  
all_data_comparison <- dplyr::select(all_data_comparison, shannon_entropy, chao1, menhinick, margalef, fisher_alpha, simpson, gini_index, strong, pielou_evenness, faith_pd, condition)

all_data_comparison$condition[all_data_comparison$condition=="healthy"] <- "AGP"

violin_hospital_compare <- vector('list', length(metric))

violin_hospital_compare <- plot_violin(all_data_comparison, "condition")

# PNG device
#jpeg("my_plot.jpeg", quality = 100)

grid.arrange(violin_hospital_compare[[1]], violin_hospital_compare[[2]], violin_hospital_compare[[3]], violin_hospital_compare[[4]], ncol=2) 

#dev.off()

grid.arrange(violin_hospital_compare[[5]], violin_hospital_compare[[6]], violin_hospital_compare[[7]], violin_hospital_compare[[8]], ncol=2) 
grid.arrange(violin_hospital_compare[[9]], violin_hospital_compare[[10]], ncol=2)

grid.arrange(violin_hospital_compare[[1]], violin_hospital_compare[[2]], violin_hospital_compare[[3]], violin_hospital_compare[[4]], violin_hospital_compare[[5]], violin_hospital_compare[[6]], violin_hospital_compare[[7]], violin_hospital_compare[[8]], violin_hospital_compare[[9]], ncol=3) 
```

```{r}
box_2 <- vector('list', length(metric))
  
for (i in 1:length(metric)){
  plot_data <- all_data_comparison %>% dplyr::group_by(condition) %>% dplyr::mutate(m = mean(.data[[metric[i]]])) 
  plot_data$condition <- factor(plot_data$condition , levels=c("CDI", "CDIpre", "CDIpost", "AGP", "donor"))

  box_2[[i]] <- plot_data %>% ggplot(aes( x = condition, y = .data[[metric[i]]], color = condition)) +
    geom_boxplot()+ #outlier.shape = NA
    ylab(label_fun(metric[i])) +
    xlab("")+
    theme_classic()+
    theme(axis.text.x = element_text(angle = 30 , hjust = 0.9))+
    font("xy.text", size = 10) +
    rremove("legend") +
    scale_color_manual(values=c('#8c510a','#bf812d','#dfc27d','#5ab4ac','#018571'))
}
#png(file = here("03_plots_and_tables", "compare_all_CDI_1.png"), width = 3000, height = 2000, res = 300)
grid.arrange(box_2[[1]], box_2[[2]], box_2[[3]], box_2[[4]], box_2[[5]], box_2[[6]], ncol=3) 
#dev.off()

#png(file = here("03_plots_and_tables", "compare_all_CDI_2.png"), width = 3000, height = 2000, res = 300)
grid.arrange(box_2[[7]], box_2[[8]], box_2[[9]], box_2[[10]], ncol=3) 
#dev.off()

```

```{r}
violin_2 <- vector('list', length(metric))
  
for (i in 1:length(metric)){
  plot_data <- all_data_comparison %>% dplyr::group_by(condition) %>% dplyr::mutate(m = mean(.data[[metric[i]]])) 

  violin_2[[i]] <- plot_data %>% ggplot(aes( x = reorder(condition, m), y = .data[[metric[i]]], color = condition, fill=condition)) +
    geom_violin()+
    geom_boxplot(color = "black", fill = "white", width=0.10,  alpha = 0.8, outlier.shape = NA) +    
    ylab(label_fun(metric[i])) +
    xlab("")+
    theme_classic()+
    font("xy.text", size = 10) +
    #rotate_x_text(30) +
    rremove("legend") +
    # scale_fill_brewer(palette="Dark2") +
    # scale_color_brewer(palette="Dark2")+
    scale_color_manual(values=c('#a6611a','#dfc27d','#80cdc1','#018571', '#d8b365'))
    
}

violin_2[[10]] + scale_y_continuous(trans='log10')

grid.arrange(violin_2[[1]], violin_2[[2]], violin_2[[3]], violin_2[[4]], ncol=2) 
grid.arrange(violin_2[[5]], violin_2[[6]], violin_2[[7]], violin_2[[8]], ncol=2) 
grid.arrange(violin_2[[9]], violin_2[[10]], ncol=2, nrow=2)

```



