---
title: "Taxonomy_analysis"
author: "Natasa Mortvanski"
date: "2023-04-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

## Taxonomy 

From tutorial: https://uw-madison-microbiome-hub.github.io/Microbiome_analysis_in-_R/
and: https://rpubs.com/mohsen/lefse_analysis_cleaned_prevalence_phyloseq

### CDI samples - before vs after FMT

```{r}
library(readr)
library(data.table)
library(stringr)
library(ggplot2)
library(plyr)
library(purrr)
library(dplyr)
library(flextable)
library(tibble)
library(ComplexHeatmap)
library(RColorBrewer)
library(ggplotify)
library(here)
library(tidyverse)
```

```{r}
#BiocManager::install("phyloseq")
#remotes::install_github("jbisanz/qiime2R")
#BiocManager::install("microbiome")
#remotes::install_github("microsud/microbiomeutilities")
library(phyloseq)
library(qiime2R)
library(microbiome)
library(microbiomeutilities)


# Importing ASVs abundance file
ASVs <- read_qza(here("00_raw_data/09_hospital_data_qiime2","table-CDI.qza"))

# Importing metadata
metadata <- read.delim(here("00_raw_data/07_hospital_CDI", "sample-metadata.tsv.gz"))
rownames(metadata) <- metadata[,1]
metadata$time <- as.numeric(metadata$time)

# Importing tree
tree <- read_qza(here("00_raw_data/09_hospital_data_qiime2","rooted-tree-CDI.qza"))
# Importing taxonomy
taxonomy <- read_qza(here("00_raw_data/09_hospital_data_qiime2","taxonomy-CDI.qza"))

taxonomy_table <- parse_taxonomy(taxonomy$data)

tax_table <- do.call(rbind, strsplit(as.character(taxonomy$data$Taxon), "; "))
colnames(tax_table) <- c("Kingdom","Phylum","Class","Order","Family","Genus","Species")
rownames(tax_table) <- taxonomy$data$Feature.ID

# Creating phyloseq object
physeq <- phyloseq(
  otu_table(ASVs$data, taxa_are_rows = TRUE),
  phy_tree(tree$data),
  tax_table(tax_table),
  sample_data(metadata)
)

#Subset the data to keep only Receptor samples.
physeq.rec <- subset_samples(physeq, sample.purpose == "Receptor")

for(i in 1:nsamples(physeq.rec)) {
  if(sample_data(physeq.rec)$time[i] <= 0){
    sample_data(physeq.rec)$FMT_pre_post[i] <- "pre"
  } else {
    sample_data(physeq.rec)$FMT_pre_post[i] <- "post"
  }
}

sample_data(physeq.rec)$FMT_pre_post <- as.factor(sample_data(physeq.rec)$FMT_pre_post)
sample_data(physeq.rec)$FMT_pre_post <- relevel(sample_data(physeq.rec)$FMT_pre_post, "pre")
levels(sample_data(physeq.rec)$FMT_pre_post)
```

```{r}
#BiocManager::install("ALDEx2")
#library(ALDEx2)

#aldex2_da <- ALDEx2::aldex(data.frame(phyloseq::otu_table(physeq.rec)), phyloseq::sample_data(physeq.rec)$FMT_pre_post, test="t", effect = TRUE, denom="iqlr")

## Vulcano plot for differentially expressed OTUs
#ALDEx2::aldex.plot(aldex2_da, type="MW", test="wilcox", called.cex = 1, cutoff = 0.1)
```


```{r}
ntaxa(physeq.rec)

nsamples(physeq.rec)

otu_table(physeq.rec)[1:5, 1:5]  

tax_table(physeq.rec)[1:5, 1:4]
```

```{r}
# Prune and rarefy samples

# Keep samples with non-zero total taxa
pruned <- prune_samples(sample_sums(physeq.rec) > 0, physeq.rec)

# Check total taxa in each sample again
sample_sums(pruned)

#rarefy samples
physeq_rarefy <- rarefy_even_depth(pruned, rngseed=1, sample.size=0.9*min(sample_sums(pruned)), replace=F)

#Check the taxa prevalence at Phylum level
#plot_taxa_prevalence(physeq_rarefy, "Order")
```

Barplots are a one way of visualising the composition of your samples. At Family level and relative abundance

```{r}
# convert to relative abundance  
physeq.rec.rel <- microbiome::transform(physeq.rec, "compositional")

physeq.rec.rel2 <- prune_taxa(taxa_sums(physeq.rec.rel) > 0, physeq.rec.rel)
```

Check for the core ASVs

```{r}
core.taxa.standard <- core_members(physeq.rec.rel2, detection = 0.001, prevalence = 50/100)
print(core.taxa.standard)
```

we only see IDs, not very informative. We can get the classification of these as below.

```{r}
#install.packages('DT')
library(DT)
# Extract the taxonomy table
taxonomy_core <- as.data.frame(tax_table(physeq.rec.rel2))

# Subset this taxonomy table to include only core OTUs
core_taxa_id <- subset(taxonomy_core, rownames(taxonomy_core) %in% core.taxa.standard)

DT::datatable(core_taxa_id)
```

```{r}
physeq.fam.rel <- physeq.rec %>%
  aggregate_rare(level = "Family", detection = 1/100, prevalence = 75/100) %>%
  microbiome::transform(transform = "compositional")

plot_composition(physeq.fam.rel, sample.sort = "FMT_pre_post", x.label = "FMT_pre_post") + 
  theme(legend.position = "bottom") + 
  scale_fill_brewer("Family", palette = "Paired") + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, size = 10)) + 
  ggtitle("Relative abundance") + 
  theme(legend.title = element_text(size = 18))

```

```{r}
library(RColorBrewer)

physeq.fam.rel <- physeq.rec %>%
  aggregate_rare(level = "Family", detection = 1/100, prevalence = 50/100) %>%
  microbiome::transform(transform = "compositional") 

colourCount = ntaxa(tax_table(physeq.fam.rel))
getPalette = colorRampPalette(brewer.pal(13, "Paired"))

plot_composition(physeq.fam.rel,
                 average_by = "FMT_pre_post", 
                 transform = "compositional") + 
  scale_fill_manual(values = getPalette(colourCount)) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, size = 10)) + 
  ggtitle("Relative abundance") + 
  theme(legend.title = element_text(size = 18))
```

```{r}
phyloseq::plot_bar(physeq.fam.rel, fill = "Family") +
  geom_bar(aes(fill = Family), stat = "identity", position = "stack") +
  labs(x = "", y = "Relative Abundance\n") +
  facet_wrap(~ FMT_pre_post, scales = "free", nrow =2) +
  theme(panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  scale_fill_manual(values = getPalette(colourCount))
```

```{r}
# install.packages(
#   "microViz",
#   repos = c(davidbarnett = "https://david-barnett.r-universe.dev", getOption("repos"))
# )
#library(microViz)
# 
# physeq.rec <- physeq.rec %>%
#  tax_fix(
#   min_length = 4,
#   unknowns = c("k__Bacteria", "p__Firmicutes", "p__Proteobacteria", "p__OD1 Phylum", "p__OD1"),
#   sep = " ", anon_unique = TRUE,
#   suffix_rank = "classified"
#  )
# 
# 
# plot_family <- comp_barplot(
#   physeq.rec.rel2,
#   "Family",
#   n_taxa = 4,
#   tax_order = sum,
#   merge_other = TRUE,
#   sample_order = "bray",
#   order_with_all_taxa = FALSE,
#   label = "SAMPLE",
#   facet_by = "condition",
#   bar_width = 1,
#   bar_outline_colour = NA,
#   #palette = distinct_palette(n_taxa),
#   tax_transform_for_ordering = "identity",
#   tax_transform_for_plot = "compositional",
#   #seriate_method = "OLO_ward",
#   keep_all_vars = TRUE,
#   #interactive = FALSE,
#   x = "SAMPLE"
# )
# 
# plot_family
```


```{r}
mycols <- c("coral", "steelblue2")

physeq.fam <- aggregate_taxa(physeq.rec, "Family")
top_f <- top_taxa(physeq.fam, 9)
top_f

top_fams <- plot_listed_taxa(physeq.fam, top_f, 
                 group= "FMT_pre_post",
                 group.colors = mycols,
                 add.violin = F,
                 dot.opacity = 0.25,
                 box.opacity = 0.25,
                 panel.arrange= "wrap")
top_fams
```


```{r}
dom.tax <- dominant_taxa(physeq.rec,level = "Family", group="FMT_pre_post")
dom.tax_pre <- dom.tax$dominant_overview %>% filter(FMT_pre_post == "pre")
dom.tax_post <- dom.tax$dominant_overview %>% filter(FMT_pre_post == "post")

head(dom.tax$dominant_overview)

head(dom.tax_pre, 10) %>%
  flextable()
```

```{r}
head(dom.tax_post, 10)%>%
  flextable()
```

```{r}
grp_abund <- get_group_abundances(physeq.rec, 
                                  level = "Family", 
                                  group="FMT_pre_post",
                                  transform = "compositional")

# clean names 
grp_abund$OTUID <- gsub("f__", "",grp_abund$OTUID)
grp_abund$OTUID <- ifelse(grp_abund$OTUID == "", 
                          "Unclassified", grp_abund$OTUID)

grp_abund_sorted <- grp_abund[order(grp_abund$mean_abundance, decreasing=TRUE),]


mean.plot_sorted <- grp_abund_sorted[1:40,] %>% # input data
  ggplot(aes(x= reorder(OTUID, mean_abundance), # reroder based on mean abundance
             y= mean_abundance,
             fill=FMT_pre_post)) + 
  geom_bar(stat = "identity", 
          position = position_dodge()) + 
  scale_fill_manual("FMT_pre_post", values=mycols) + # manually specify colors
  theme_bw() +
  ylab("Mean Relative Abundance") + 
  xlab("Family") + 
  coord_flip() 
# +
#   scale_fill_manual(values = getPalette(3)) 

mean.plot_sorted
```

Tree plot

```{r}
physeq_top_50 <- subset_taxa(physeq.rec, Kingdom=="k__Bacteria")
physeq_top_50 <- prune_taxa(names(sort(taxa_sums(physeq_top_50),TRUE)[1:50]), physeq_top_50)

# Color the nodes by category
plot_tree(physeq_top_50, nodelabf=nodeplotblank, label.tips="Genus", ladderize="left", color="FMT_pre_post")

# Convert to radial tree
plot_tree(physeq_top_50, nodelabf=nodeplotblank, label.tips="Genus", ladderize="left", color="FMT_pre_post") + coord_polar(theta="y")
```

Lefse to test for differential abundance between categories

from: https://rpubs.com/mohsen/lefse_analysis_cleaned_prevalence_phyloseq

```{r}
# Install microbiome marker package
#remotes::install_version('ggplot2', version='3.3.6')
#BiocManager::install("ggtree")
#BiocManager::install("microbiomeMarker")

# activate microbiome marker package 
library(microbiomeMarker)

# Differential abundance using lefse
lefse <- run_lefse(physeq.rec,
                   group = "FMT_pre_post",
                   taxa_rank = "Family",
                   wilcoxon_cutoff = 0.05,
                   norm = "CPM",
                   kw_cutoff = 0.05,
                   multigrp_strat = TRUE,
                   lda_cutoff = 2
)


#head(marker_table(lefse))

# bar plot
plot_ef_bar(lefse)
# dot plot
plot_ef_dot(lefse)

```

```{r}
library(knitr)
dat <- marker_table(lefse) %>% data.frame() %>% select(1:4)

dat %>% flextable()
```


## Lefse with lefser package

How to transform data in SE format: https://microbiome.github.io/OMA/data-introduction.html

```{r}
# #BiocManager::install("lefser")
# library(lefser)
# #BiocManager::install("microbiome/mia", version='3.14')
# library(mia)
# 
# counts  <- ASVs$data   # Abundance table (e.g. ASV data; to assay data)
# tax     <- tax_table     # Taxonomy table (to rowData)
# samples <-  metadata  # Sample data (to colData)
# se <- SummarizedExperiment(assays = list(counts = counts),
#                            colData = samples,
#                            rowData = tax)
# 
# # Goes through the whole DataFrame. Removes '.*[kpcofg]__' from strings, where [kpcofg] 
# # is any character from listed ones, and .* any character.
# rowdata_modified <- BiocParallel::bplapply(rowData(se), 
#                                            FUN = stringr::str_remove, 
#                                            pattern = '.*[kpcofg]__')
# 
# # Genus level has additional '\"', so let's delete that also
# rowdata_modified <- BiocParallel::bplapply(rowdata_modified, 
#                                            FUN = stringr::str_remove, 
#                                            pattern = '\"')
# 
# # rowdata_modified is a list, so it is converted back to DataFrame format. 
# rowdata_modified <- DataFrame(rowdata_modified)
# 
# # And then assigned back to the SE object
# rowData(se) <- rowdata_modified
# 
# # Now we have a nicer table
# head(rowData(se))
# 
# head(colData(se))
# 
# #add relative abundances along the original count data 
# se <- relAbundanceCounts(se)
# assays(se)
# 
# assays(se)$counts[1:5,1:7]
# assay(se, "relabundance")[1:5,1:7]
# 
# dim(se)
# 
# #add phylogenetic tree
# tse <- as(se, "TreeSummarizedExperiment")
# # Add tree to rowTree
# rowTree(tse) <- tree$data
# # Check
# tse
# 
# se <- tse
# 
# #head(rowTree(tse))
# rowLinks(tse)
# 
# # subset by sample purpose
# se_subset_by_sample <- se[ , se$sample.purpose %in% c("Receptor")]
# 
# # show dimensions
# dim(se_subset_by_sample)
# 
# head(colData(se_subset_by_sample))
# 
# # add metadata column
# 
# for(i in 1:dim(se_subset_by_sample)[2]) {
#   if(se_subset_by_sample$time[i] <= 0){
#     se_subset_by_sample$FMT_pre_post[i] <- "pre"
#   } else {
#     se_subset_by_sample$FMT_pre_post[i] <- "post"
#   }
# }
# 
# 
# # subset by sample and feature and get rid of NAs
# se_subset_by_sample_feature <- se[rowData(se)$Phylum %in% c("Bacteria") & !is.na(rowData(se)$Phylum), se$sample.purpose %in% c("Receptor")]
# 
# # show dimensions
# dim(se_subset_by_sample_feature)
```

```{r}
# table(se_subset_by_sample$FMT_pre_post)
```

```{r}
# res <- lefser(se_subset_by_sample, groupCol = "FMT_pre_post")
# head(res)
# lefserPlot(res)
```

### Donor data vs CDI data

```{r}
# Importing ASVs abundance file
ASVs_merged <- read_qza(here("00_raw_data/09_hospital_data_qiime2","merged-table.qza"))

# Importing metadata
metadata_CDI <- read.delim(here("00_raw_data/07_hospital_CDI", "sample-metadata.tsv.gz"))
metadata_CDI$time <- as.numeric(metadata_CDI$time)
names(metadata_CDI)[names(metadata_CDI) == 'Sample.ID'] <- 'SampleID'
names(metadata_CDI)[names(metadata_CDI) == 'sample.origin'] <- 'condition'


for(i in 1:nrow(metadata_CDI)) {
  if(metadata_CDI$time[i] <= 0 & !is.na(metadata_CDI$time[i])){
    metadata_CDI$condition[i] <- paste(metadata_CDI$condition[i], "pre", sep="_")
  } else if (metadata_CDI$time[i] > 0 & !is.na(metadata_CDI$time[i])) {
    metadata_CDI$condition[i] <- paste(metadata_CDI$condition[i], "post", sep="_")
  }
}

metadata_donor <- read.delim(here("00_raw_data/08_hospital_donor","sample-metadata.tsv.gz"))
metadata_donor$condition <- "healthy_donors"
names(metadata_donor)[names(metadata_donor) == 'sample_id'] <- 'SampleID'


metadata_merged <- rbind.fill(metadata_CDI, metadata_donor)
metadata_merged <- dplyr::select(metadata_merged, SampleID, condition)
rownames(metadata_merged) <- metadata_merged[,1]


# Importing tree
tree_merged <- read_qza(here("00_raw_data/09_hospital_data_qiime2","rooted-tree-merged.qza"))

# Importing taxonomy
taxonomy_merged <- read_qza(here("00_raw_data/09_hospital_data_qiime2","taxonomy-merged.qza"))
taxonomy_table_merged <- parse_taxonomy(taxonomy_merged$data)
tax_table_merged <- do.call(rbind, strsplit(as.character(taxonomy_merged$data$Taxon), "; "))
colnames(tax_table_merged) <- c("Kingdom","Phylum","Class","Order","Family","Genus","Species")
rownames(tax_table_merged) <- taxonomy_merged$data$Feature.ID

# Creating phyloseq object
physeq_merged <- phyloseq(
  otu_table(ASVs_merged$data, taxa_are_rows = TRUE),
  phy_tree(tree_merged$data),
  tax_table(tax_table_merged),
  sample_data(metadata_merged)
)

#Subset the data to keep only Receptor samples.
physeq_merged <- subset_samples(physeq_merged, condition == c("Cdif_pre", "Cdif_post", "healthy_donors"))

sample_data(physeq_merged)$condition <- as.factor(sample_data(physeq_merged)$condition)
sample_data(physeq_merged)$condition <- relevel(sample_data(physeq_merged)$condition, "Cdif_pre")
levels(sample_data(physeq_merged)$condition)
```

Barplots :

```{r}
# convert to relative abundance  
physeq.rel <- microbiome::transform(physeq_merged, "compositional")

physeq.rel2 <- prune_taxa(taxa_sums(physeq.rel) > 0, physeq.rel)
```


```{r}
physeq.fam.rel <- physeq_merged %>%
  aggregate_rare(level = "Family", detection = 1/100, prevalence = 75/100) %>%
  microbiome::transform(transform = "compositional")

plot_composition(physeq.fam.rel, sample.sort = "condition", x.label = "condition") + 
  theme(legend.position = "bottom") + 
  scale_fill_brewer("Family", palette = "Paired") + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, size = 10)) + 
  ggtitle("Relative abundance") + 
  theme(legend.title = element_text(size = 18))
```

```{r}
 library(RColorBrewer)

physeq.fam.rel <- physeq_merged %>%
  aggregate_rare(level = "Family", detection = 1/100, prevalence = 50/100) %>%
  microbiome::transform(transform = "compositional") 

colourCount = ntaxa(tax_table(physeq.fam.rel))
getPalette = colorRampPalette(brewer.pal(13, "Paired"))

plot_composition(physeq.fam.rel,
                 average_by = "condition", 
                 transform = "compositional") + 
  scale_fill_manual(values = getPalette(colourCount)) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, size = 10)) + 
  ggtitle("Relative abundance") + 
  theme(legend.title = element_text(size = 18))
```

```{r}
mycols <- c("coral", "steelblue2", "darkolivegreen2")

physeq.fam <- aggregate_taxa(physeq_merged, "Family")

sample_data(physeq.fam)$condition <- as.factor(str_replace(sample_data(physeq.fam)$condition, "Cdif_pre", "Pre"))
sample_data(physeq.fam)$condition <- as.factor(str_replace(sample_data(physeq.fam)$condition, "Cdif_post", "Post"))
sample_data(physeq.fam)$condition <- as.factor(str_replace(sample_data(physeq.fam)$condition, "healthy_donors", "Donors"))

sample_data(physeq.fam)$condition <- relevel(sample_data(physeq.fam)$condition, ref = "Post")
sample_data(physeq.fam)$condition <- relevel(sample_data(physeq.fam)$condition, ref = "Pre")

top_f <- top_taxa(physeq.fam, 9)
top_f

top_fams <- plot_listed_taxa(physeq.fam,
                             top_f,
                             group= "condition",
                             group.colors = mycols,
                             add.violin = F,
                             dot.opacity = 0.25,
                             box.opacity = 0.25,
                             panel.arrange= "wrap")
top_fams
```

```{r}
dom.tax <- dominant_taxa(physeq_merged, level = "Family", group="condition")
dom.tax_pre <- dom.tax$dominant_overview %>% filter(condition == "Cdif_pre")
dom.tax_post <- dom.tax$dominant_overview %>% filter(condition == "Cdif_post")
dom.tax_donor <- dom.tax$dominant_overview %>% filter(condition == "healthy_donors")

head(dom.tax_pre) %>%
  flextable()
```

```{r}
head(dom.tax_post)%>%
  flextable()
```

```{r}
head(dom.tax_donor)%>%
  flextable()
```


```{r}
grp_abund <- get_group_abundances(physeq_merged, 
                                  level = "Family", 
                                  group="condition",
                                  transform = "compositional")

# clean names 
grp_abund$OTUID <- gsub("f__", "",grp_abund$OTUID)
grp_abund$OTUID <- ifelse(grp_abund$OTUID == "", 
                          "Unclassified", grp_abund$OTUID)

grp_abund_sorted <- grp_abund[order(grp_abund$mean_abundance, decreasing=TRUE),]


mean.plot_sorted <- grp_abund_sorted[1:40,] %>% # input data
  ggplot(aes(x= reorder(OTUID, mean_abundance), # reroder based on mean abundance
             y= mean_abundance,
             fill=condition)) + 
  geom_bar(stat = "identity", 
          position = position_dodge()) + 
  scale_fill_manual("condition", values=mycols) + # manually specify colors
  theme_bw() +
  ylab("Mean Relative Abundance") + 
  xlab("Family") + 
  coord_flip()

mean.plot_sorted
```

Tree plot

```{r}
physeq_top_50 <- subset_taxa(physeq_merged, Kingdom=="k__Bacteria")
physeq_top_50 <- prune_taxa(names(sort(taxa_sums(physeq_top_50),TRUE)[1:50]), physeq_top_50)

# Color the nodes by category
plot_tree(physeq_top_50, nodelabf=nodeplotblank, label.tips="Genus", ladderize="left", color="condition")

# Convert to radial tree
plot_tree(physeq_top_50, nodelabf=nodeplotblank, label.tips="Genus", ladderize="left", color="condition") + coord_polar(theta="y")
```

Lefse to test for differential abundance between categories

from: https://rpubs.com/mohsen/lefse_analysis_cleaned_prevalence_phyloseq

```{r}
# Differential abundance using lefse
lefse <- run_lefse(physeq_merged,
                   group = "condition",
                   taxa_rank = "Family",
                   wilcoxon_cutoff = 0.05,
                   norm = "CPM",
                   kw_cutoff = 0.05,
                   multigrp_strat = TRUE,
                   lda_cutoff = 2
)

# bar plot
plot_ef_bar(lefse)
# dot plot
plot_ef_dot(lefse)
```

```{r}
library(knitr)
dat <- marker_table(lefse) %>% data.frame() %>% select(1:4)

dat %>% flextable()
```
