---
title: "Taxonomy_correlation"
author: "Natasa Mortvanski"
date: "2023-11-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(here)
library(phyloseq)
library(qiime2R)
library(microbiome)
library(microbiomeutilities)
library(tidyr)
library(data.table)
library(corrplot)
library(flextable)

# Creating phyloseq object

physeq <- qza_to_phyloseq(
  features= "~/Desktop/master_project/Data_Hospital_donors/R2960/SCFA_QIIME2_analysis/table-dada2.qza",
  tree= "~/Desktop/master_project/Data_Hospital_donors/R2960/SCFA_QIIME2_analysis/rooted-tree.qza",
  taxonomy= "~/Desktop/master_project/Data_Hospital_donors/R2960/SCFA_QIIME2_analysis/taxonomy.qza",
  metadata = "~/Desktop/master_project/Master-Project-Natasa-Mortvanski/01_tidy_data/SCFA_alpha_donors.tsv"
  )

physeq@tax_table
```

## Taxonomic level: Family

### Creating a table of abundances on Family level:

```{r}
# tax_glom function merges species that have the same taxonomy at a certain 
# taxonomic rank - in this case family level
glom <- tax_glom(physeq, taxrank=rank_names(physeq)[5])

# psmelt function melts phyloseq data object into large data.frame 
smelted_data <- psmelt(glom) 

abundances_fam <- smelted_data %>% select(-c(Kingdom, Phylum, Class, Order)) %>%
        spread(Family, Abundance) %>%
  group_by(Sample) %>%
  summarize(across(3:63, ~unique(na.omit(.x))),.groups = 'drop')

abundances_fam
```

### First lets check correlations between alpha diversity metrics and family abundances:

```{r}
# Save results as data frame
tax_alpha_cor <- as.data.frame(cor(abundances_fam[, c(6:15)], abundances_fam[,c(16:50)]))
tax_alpha_cor %>% tibble::rownames_to_column() %>% flextable()

# Plot correlation 
matrix_tax_alpha <- cor(abundances_fam[, c(6:15)], abundances_fam[,c(16:50)] , method = "spearman")
corrplot(matrix_tax_alpha, method="circle", tl.col = "black", tl.srt = 90, col.lim = c(-1,1), is.corr = FALSE)
```

### Then lets check correlations between SCFA and family abundances:

```{r}
library(dichromat)
library(RColorBrewer)
library(graphics)


# Save results as data frame
tax_SCFA_cor <- as.data.frame(cor(abundances_fam[, c(3:5)], abundances_fam[,c(16:50)]))
tax_SCFA_cor %>% tibble::rownames_to_column() %>% flextable()


scalebluered <- colorRampPalette(brewer.pal(8, "RdBu"))(50)

# Plot correlation 
matrix_tax_SCFA <- cor(abundances_fam[, c(3:5)], abundances_fam[,c(16:50)] , method = "spearman")

#corrplot(matrix_tax_SCFA, method="circle", tl.col = "black", tl.srt = 90, col.lim = c(-1,1), is.corr = FALSE) 

# I did not like how the generic legend looked like, so I decided to generate legend separately. However,
# I did not know how to put it on right side of the plot to look like the other plots, so I put it on the left side...

corrplot(matrix_tax_SCFA, method="circle", tl.col = "black", tl.srt = 90, cl.pos="n") 
colorlegend(xlim=c(-3,0), ylim=c(10,5), scalebluered, c(seq(-1,1,1)), align="r", vertical=TRUE, addlabels=TRUE)
```

### Finally, lets check correlations between alpha diversity metrics and SCFA:

```{r}
# Save results as data frame
alpha_SCFA_cor <- as.data.frame(cor(abundances_fam[, c(3:5)], abundances_fam[,c(6:15)]))
alpha_SCFA_cor %>% tibble::rownames_to_column() %>% flextable()

# Plot correlation 
matrix_alpha_SCFA <- cor(abundances_fam[, c(3:5)], abundances_fam[,c(6:15)] , method = "spearman")
corrplot(matrix_alpha_SCFA, method="circle", tl.col = "black", tl.srt = 90, col.lim = c(-1,1), is.corr = FALSE)
```

### Correlation plot for family abundancies vs. SCFA and alpha diversity all together:

```{r}
matrix_all <- cor(abundances_fam[, c(3:15)], abundances_fam[,c(16:50)] , method = "spearman")
corrplot(matrix_all, method="circle", tl.col = "black", tl.srt = 90, col.lim = c(-1,1), is.corr = FALSE)
```

## Taxonomic level: Genus

### Creating a table of abundances on Genus level:
 
```{r}
# tax_glom function merges species that have the same taxonomy at a certain 
# taxonomic rank - in this case genus level
glom <- tax_glom(physeq, taxrank=rank_names(physeq)[6])

# psmelt function melts phyloseq data object into large data.frame 
smelted_data <- psmelt(glom) 

# Problem with Clostridium!
abundances_gen <- smelted_data %>% select(-c(Kingdom, Phylum, Class, Order, Family)) %>%
        spread(Genus, Abundance) %>%
  group_by(Sample) %>% select(-c(Clostridium)) %>%
  summarize(across(2:94, ~unique(na.omit(.x))),.groups = 'drop')

abundances_gen
```

### First lets check correlations between alpha diversity metrics and family abundances:

```{r}
# Save results as data frame
tax_alpha_cor <- as.data.frame(cor(abundances_gen[, c(7:16)], abundances_gen[,c(17:94)]))
tax_alpha_cor %>% tibble::rownames_to_column() %>% flextable()

# Plot correlation 
matrix_tax_alpha_1 <- cor(abundances_gen[, c(7:16)], abundances_gen[,c(17:55)] , method = "spearman")
corrplot(matrix_tax_alpha_1, method="circle", tl.col = "black", tl.srt = 90, tl.cex =0.9, col.lim = c(-1,1), is.corr = FALSE)

matrix_tax_alpha_2 <- cor(abundances_gen[, c(7:16)], abundances_gen[,c(56:94)] , method = "spearman")
corrplot(matrix_tax_alpha_2, method="circle", tl.col = "black", tl.srt = 90, tl.cex =0.9, col.lim = c(-1,1), is.corr = FALSE)
```

### Then lets check correlations between SCFA and family abundances:

```{r}
# Save results as data frame
tax_SCFA_cor <- as.data.frame(cor(abundances_gen[, c(4:6)], abundances_gen[,c(17:94)]))
tax_SCFA_cor %>% tibble::rownames_to_column() %>% flextable()

# Plot correlation 
matrix_tax_SCFA_1 <- cor(abundances_gen[, c(4:6)], abundances_gen[,c(17:55)] , method = "spearman")
corrplot(matrix_tax_SCFA_1, method="circle", tl.col = "black", tl.srt = 90, cl.pos="n", is.corr = FALSE)
colorlegend(xlim=c(-3,0), ylim=c(10,5), scalebluered, c(seq(-1,1,1)), align="r", vertical=TRUE, addlabels=TRUE)

matrix_tax_SCFA_2 <- cor(abundances_gen[, c(4:6)], abundances_gen[,c(56:94)] , method = "spearman")
corrplot(matrix_tax_SCFA_2, method="circle", tl.col = "black", tl.srt = 90, cl.pos="n", is.corr = FALSE)
colorlegend(xlim=c(-3,0), ylim=c(10,5), scalebluered, c(seq(-1,1,1)), align="r", vertical=TRUE, addlabels=TRUE)

```

### Correlation plot for genus abundancies vs. SCFA and alpha diversity all together:

```{r}
matrix_all_1 <- cor(abundances_gen[, c(4:16)], abundances_gen[,c(17:55)] , method = "spearman")
corrplot(matrix_all_1, method="circle", tl.col = "black", tl.srt = 90, tl.cex =0.9, col.lim = c(-1,1), is.corr = FALSE)

matrix_all_2 <- cor(abundances_gen[, c(4:16)], abundances_gen[,c(56:94)] , method = "spearman")
corrplot(matrix_all_2, method="circle", tl.col = "black", tl.srt = 90, tl.cex =0.9, col.lim = c(-1,1), is.corr = FALSE)
```
